# Data Import - MAG Catalogue

## Overview

**WARNING: This is the MAG catalogue from short-read assemblies and not hybrid assemblies with PacBio.**

This chapter prepares the Metagenome-Assembled Genome (MAG) catalogue for downstream analyses. All data processing steps are shown for transparency, but code chunks are set to `eval=FALSE` so they do not execute during bookdown compilation. The processed data is saved to `.Rdata` files which are then loaded in subsequent plotting chapters.

## Genome (MAGs) Catalogue

Relevant metadata of genomes are fetched and merged into one genome metadata object for downstream analyses. This includes taxonomic classification, quality metrics, phylogenetic trees, and functional annotations.

Location on ERDA:
ID: ELPE6LATFN
Path: 3D-omics/processed_data/f_mg-mg_assembly
Access: read

genome_taxonomy: https://sid.erda.dk/share_redirect/ELPE6LATFN/results/prokaryotes/annotate/gtdbtk.summary.tsv
genome_quality: https://sid.erda.dk/share_redirect/ELPE6LATFN/results/prokaryotes/annotate/checkm2.quality_report.tsv
genome_tree: https://sid.erda.dk/share_redirect/ELPE6LATFN/results/prokaryotes/annotate/gtdbtk.backbone.bac120.classify.tree
genome_annotations: https://sid.erda.dk/share_redirect/ELPE6LATFN/results/prokaryotes/annotate/dram.annotations.tsv.gz

The number of genomes in these files should match the number of genomes in the count tables, after accounting for the appropriate dereplication at species level.
I find the drep.0.95.fa.gz file on ERDA.
To find which genomes exist at 0.95 dereplication level I do the following in the terminal:

cd OneDrive - University of Copenhagen 
cd Documents/Github\ projects/Adenovirus/data/MAG_catalogue 
gunzip drep.0.95.fa.gz
grep "^>" drep.0.95.fa | sed 's/>//' > genome_names.tsv 
cut -d'@' -f1 genome_names.tsv | sort -u > unique_genome_names.tsv


### Dereplicated Genomes

Bins were initially dereplicated at 99%, and annotations generated for all resulting MAGs. A second dereplication was performed at 95% to yield species-representative genomes. This analysis is performed on that subset.

**Note:** The `drep.0.95.fa` file must be processed first using terminal commands (shown in comments above) to extract unique genome names.

```{r dereplicated_genomes_01A, message=FALSE, warning=FALSE, eval=FALSE}
# Read list of genomes at 95% dereplication level
# This file is generated from the drep.0.95.fa file using terminal commands
# Note: warning=FALSE suppresses vroom parsing warnings (e.g., inconsistent column types)
selected_genomes <- read_tsv("data/MAG_catalogue/unique_genome_names.tsv", col_names = "genome") 

# Report number of species-representative genomes
nrow(selected_genomes)
```

The dereplication yielded 251 species-representative genomes that will be used for all downstream analyses.



### Genome (MAGs) Taxonomy

This section processes the raw taxonomy table generated by GTDBtk and simplifies it for downstream analyses. Taxonomic classifications are parsed, cleaned, and formatted consistently.

```{r load_taxonomy_01A, message=FALSE, warning=FALSE, eval=FALSE}
# Load and process GTDBtk taxonomy summary file
# This contains taxonomic classifications for all MAGs at species level dereplication
# Note: Some rows have "N/A" in numeric columns (msa_percent, translation_table), 
# which causes parsing warnings. We suppress these and handle them properly.
genome_taxonomy <- read_tsv(
  "data/MAG_catalogue/gtdbtk.summary.tsv", 
  show_col_types = FALSE,
  na = c("", "N/A", "NA")  # Treat "N/A" as missing values
) %>%
  # Rename first column to standardize naming
  rename(genome = user_genome) %>% 
  # Remove .fa extension from genome names for consistency
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>% 
  # Filter to only include species-representative genomes (95% dereplication)
  filter(genome %in% selected_genomes$genome) %>%
  # Parse the classification string into separate taxonomic level columns
  # GTDBtk format: d__Domain;p__Phylum;c__Class;o__Order;f__Family;g__Genus;s__Species
  separate(classification, c("domain", "phylum", "class", "order", "family", 
                             "genus", "species"), sep = ";") %>% 
  # Remove GTDBtk prefix letters (d__, p__, etc.) from each taxonomic level
  mutate(across(domain:species, ~ str_sub(., 4))) %>% 
  # Select relevant taxonomic columns for downstream use
  select(genome, domain, phylum, class, order, family, genus, species) %>%
  # Handle missing or unclassified species names
  mutate(
    species = if_else(
      is.na(species) | species == "",
      paste0("unclassified_", genus),  # Use genus name if species missing
      species
    )
  ) %>%
  # Handle cases where both species and genus are unclassified
  mutate(
    species = if_else(
      is.na(species) | species == "unclassified_",
      paste0("unclassified_", family),  # Use family name if both missing
      species
    )
  ) %>%
  # Handle missing genus names
  mutate(
    genus = if_else(
      is.na(genus) | genus == "",
      paste0("unclassified_", family),
      genus
    )
  ) %>%
  # Handle duplicate species names by adding suffixes (_sp1, _sp2, etc.)
  group_by(species) %>%
  mutate(
    species = if (n() > 1) paste0(species, "_sp", row_number()) else species
  ) %>%
  ungroup()

# Rename Bacillota_I to Bacillota for consistency with other MAG catalogues
genome_taxonomy <- genome_taxonomy %>%
  mutate(phylum = case_when(
    phylum == "Bacillota_I" ~ "Bacillota",
    TRUE ~ phylum  # Keep all other phylum values as-is
  ))
```

### Genome (MAGs) Quality

Quality properties of the genomes are assessed using CheckM2, which provides completeness, contamination, and other quality metrics for each MAG.

```{r load_quality_01A, message=FALSE, warning=FALSE, eval=FALSE}
# Load CheckM2 quality report
# This contains completeness, contamination, genome size, and other quality metrics
# Note: warning=FALSE suppresses vroom parsing warnings (e.g., inconsistent column types)
genome_quality <- read_tsv("data/MAG_catalogue/checkm2.quality_report.tsv", show_col_types = FALSE) %>%
  # Rename first column (genome name column)
  rename(genome = 1) %>% 
  # Remove .fa extension from genome names for consistency
  mutate(genome = str_replace_all(genome, "\\.fa", "")) %>%
  # Filter to only include species-representative genomes (95% dereplication)
  filter(genome %in% selected_genomes$genome) %>%   
  # Select and rename relevant quality metric columns
  select(genome, Completeness, Contamination, Coding_Density, Genome_Size, Contig_N50) %>% 
  rename(completeness = Completeness, 
         contamination = Contamination, 
         coding_density = Coding_Density, 
         length = Genome_Size,
         N50_length = Contig_N50) %>%
  # Calculate circularity indicator: N50/length ratio
  # If N50 >= length, the genome is likely circularized
  mutate(circularity = N50_length/length) %>%
  mutate(circularity = ifelse(circularity >= 1, "1", "0")) %>%
  # For circularized genomes, assume 100% completeness and 0% contamination
  mutate(completeness = ifelse(circularity == "1", 100, completeness),
         contamination = ifelse(circularity == "1", 0, contamination)) %>%   
  # Calculate length ratio relative to median bacterial genome length
  # This helps identify unusually large or small genomes
  mutate(length_ratio = length/median(length, na.rm = TRUE))

# Report number of circularized genomes
cat("Number of circularized genomes:", sum(genome_quality$circularity == "1"), "\n")
```


### Genome (MAGs) Phylogenetic Tree

This section loads and processes the phylogenetic tree generated by GTDBtk. The tree is pruned to include only the species-representative genomes used in this analysis. Note that the archaeal tree is only generated if archaea are detected among the genomes.

```{r load_tree_01A, message=FALSE, warning=FALSE, eval=FALSE}
# Load GTDBtk phylogenetic tree (bacterial backbone tree)
genome_tree <- read.tree("data/MAG_catalogue/gtdbtk.backbone.bac120.classify.tree")

# Remove single quotes from tip labels (genome names) for consistency
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label, "'", "") 

# Prune tree to keep only tips corresponding to selected genomes (95% dereplication)
genome_tree <- keep.tip(genome_tree, tip = selected_genomes$genome)
```

**Note:** If there are unmatched tip labels between the tree and selected genomes, this indicates some genome names don't match between the tree file and the dereplication list. This may need manual checking.

It was giving error:
Error in keep.tip.phylo(genome_tree, tip = selected_genomes$genome) : 
  unmatched tip labels:
TH4_04:bin_000018
I will remove this taxon.

### Merged Genome (MAGs) taxonomy & quality

Merge taxonomy, length and quality information

```{r create_genomemetadata, message=FALSE, eval=FALSE}
genome_metadata <- genome_taxonomy %>%
  left_join(genome_quality, by = join_by(genome == genome))

genome_metadata$order <- gsub("-", "_", genome_metadata$order)
genome_metadata <- genome_metadata[match(genome_tree$tip.label, genome_metadata$genome), ]
```

#### Factoring of taxonomic levels
```{r phylum_order_metadata_01A, message=FALSE, warning=FALSE, eval=FALSE}

phylum_metadata <- genome_metadata %>%
  group_by(phylum) %>%  
  summarize(
    # Count the microsamples
    `number_of_genera` = n_distinct(genus),  
    # Collect unique values separated by commas
    genus = paste(unique(genus), collapse = ", "),
    phylum = paste(unique(phylum), collapse = ", "),
    .groups = "drop") %>%
  select(phylum, number_of_genera, genus) %>%
  # Preserve original order
  arrange(match(phylum, unique(genome_metadata$phylum)))  


order_metadata <- genome_metadata %>%
  group_by(order) %>%  
  summarize(
    # Count the microsamples
    `number_of_genera` = n_distinct(genus),  
    # Collect unique values separated by commas
    genus = paste(unique(genus), collapse = ", "),
    phylum = paste(unique(phylum), collapse = ", "),
    class = paste(unique(class), collapse = ", "),
    # Ensure the dataframe is ungrouped
    .groups = "drop") %>%
  select(phylum, class, order, number_of_genera, genus) %>%
  # Preserve original order
  arrange(match(order, unique(genome_metadata$order)))  

phylum_level_vector <- as.character(unique(phylum_metadata$phylum))
order_level_vector <- as.character(unique(order_metadata$order))
```



### Genome (MAGs) functional annotations

This is the raw annotation table generated by DRAM, which is used to generate GIFT data using distillR.

```{r load_annotations_01A, eval=FALSE, message=FALSE, warning=FALSE}
# Load DRAM functional annotations
# Note: warning=FALSE suppresses vroom parsing warnings (e.g., inconsistent column types)
genome_annotations <- read_tsv("data/MAG_catalogue/dram.annotations.tsv.gz", show_col_types = FALSE) %>%
  rename(gene = 1, genome = 2) 

genome_annotations <- genome_annotations %>% filter(genome %in% selected_genomes$genome)
```


### Distil functional annotations

Raw functional annotations are distilled into genome-inferred functional traits to generate biologically more meaningful functional traits for downstream analyses.

```{r distill_annotations, warning=FALSE, comments="", message=FALSE, results='hide', eval=FALSE}
genome_gifts <- distill(genome_annotations, GIFT_db, genomecol = 2, annotcol = c(9, 10, 19))
```


## Bacterial phyla & order color scheme data
```{r set_taxa_colours, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- c(Actinomycetota = "#346254",
                   Bacillota =  "#4a6ab7",
                   Bacillota_A = "#8c1c47",
                   Bacillota_B = "#644ca3",
                   Bacillota_C = "#144f5c",
                   Bacillota_G = "#990000",
                   Bacteroidota = "#9c8464",
                   Cyanobacteriota = "#b45f06",
                   Desulfobacterota = "#462410",
                   Pseudomonadota = "#c49d4b"
                   )

order_colors <- c(Actinomycetales	= "#79b1a3",
                  Coriobacteriales =	"#498a77",
                  
                  Acholeplasmatales      = "#cfe2f3",
                  CAJFEE01 =	"#a0c5e8",
                  Erysipelotrichales =	"#7acef4",
                  Lactobacillales =	"#6390fb",
                  ML615J_28 =	"#70c9de",
                  RF39 =	"#5c7fba",
                  RFN20                  = "#6390bb",
                  MOL361        = "#70a8dc",
                  
              
                  Christensenellales =	"#f58262",
                  Clostridiales =	"#cf4a82",
                  Lachnospirales =	"#e15f7d",
                  Monoglobales = "#e18299",
                  Oscillospirales =	"#ba2760",
                  Peptostreptococcales =	"#e35d51",
                  TANB77 =	"#ef3d26",
                  UBA1381 =	"#be5643",
                  UBA1212      = "#cc0100",
                  HGM11327 = "#d04c68",
                  Eubacteriales = "#c93232", 
                    
                  Peptococcales       = "#d9d2e9",
                  UBA4068        = "#b4a7d6",
                  UBA7702      = "#8e7cc3",
                  
                  Veillonellales = "#2c7c91",
                  
                  DTUO25 = "#cc1a1a",
                 
                  Gastranaerophilales    = "#e69138",
                  Desulfovibrionales = "#8b4e2e",
                  
                  Bacteroidales =	"#e3c7a0",
                  Enterobacterales =	"#ffcc62",
                  RF32 = "#f2d78c",
                  Rs_D84 = "#f2c98e",
                  Verrucomicrobiales = "#823f1d")
```




## Save Working Objects

In the last step, the objects needed for downstream analyses are saved to `.Rdata` files. Objects are split into logical groups to keep file sizes manageable for GitHub.

### Save Core MAG Catalogue Data (for plotting chapters)

```{r wrap_objects_core, eval=FALSE}
# Save core MAG catalogue objects used in plotting chapters
save(
  genome_tree,
  genome_metadata,
  phylum_level_vector,
  order_level_vector,
  genome_gifts,
  file = "data/MAG_catalogue/data.Rdata"
)
```

### Save Color Schemes (for plotting chapters)

```{r wrap_objects_colors, eval=FALSE}
# Save color schemes and plotting theme for downstream plots
# Note: treatment_colours_bright and custom_ggplot_theme will be added in 01B
save(
  phylum_colors,
  order_colors,
  file = "data/data_colors.Rdata"
)
```

### Save Complete Preparation Data (for colleagues' analyses)

For colleagues who need access to all preparation data for additional analyses or plotting, the data is saved in multiple smaller files organized by data type:

```{r wrap_objects_taxonomy_quality, eval=FALSE}
# Save genome selection, taxonomy, and quality data
# This file contains basic genome information needed for custom analyses
save(
  selected_genomes,      # List of species-representative genomes (95% dereplication)
  genome_taxonomy,        # Full taxonomic classification for all genomes
  genome_quality,         # Quality metrics (completeness, contamination, etc.)
  file = "data/MAG_catalogue/prep_taxonomy_quality.Rdata"
)
```

```{r wrap_objects_metadata_summaries, eval=FALSE}
# Save taxonomic metadata summaries
# Contains aggregated taxonomic information (phylum and order level summaries)
save(
  phylum_metadata,        # Summary statistics at phylum level
  order_metadata,         # Summary statistics at order level
  file = "data/MAG_catalogue/prep_metadata_summaries.Rdata"
)
```

```{r wrap_objects_annotations, eval=FALSE}
# Save genome annotations (functional annotations from DRAM)
# This is likely the largest file as it contains detailed functional annotations
save(
  genome_annotations,    # Functional annotations for all genomes (DRAM output)
  file = "data/MAG_catalogue/prep_annotations.Rdata"
)
```

```{r wrap_objects_complete_backup, eval=FALSE}
# Optional: Save all objects together for convenience
# This is a complete backup that includes all objects from preparation steps
# Includes duplicates of core objects for self-contained access
save(
  selected_genomes,
  genome_taxonomy,
  genome_quality,
  genome_tree,            # Also in data/MAG_catalogue/data.Rdata
  genome_metadata,        # Also in data/MAG_catalogue/data.Rdata
  phylum_metadata,
  order_metadata,
  phylum_level_vector,    # Also in data/MAG_catalogue/data.Rdata
  order_level_vector,     # Also in data/MAG_catalogue/data.Rdata
  genome_annotations,
  genome_gifts,           # Also in data/MAG_catalogue/data.Rdata
  phylum_colors,          # Also in data/data_colors.Rdata
  order_colors,           # Also in data/data_colors.Rdata
  file = "data/01_data_preparation_complete.Rdata"
)
```

- **read_counts**: Number of reads mapped to each genome in each sample. Note this is the unfiltered and unnormalised raw community composition table.
- **genome_counts**: Number of genomes quantified in each sample, calculated through filtering and normalising ***read_counts***. This is the community composition table to be used in downstream analyses unless otherwise stated.
- **genome_tree**: Phylogenetic tree of the genomes, to be employed in downstream phylogenetic analyses.
- **genome_metadata**: Taxonomic and quality information of the genomes.
- **sample_metadata**: Treatment/population and other relevant metadata of the samples.
