# Data Preparation for Plotting

## Overview

This chapter transforms the processed metadata and sequencing statistics into tidy data frames optimized for plotting. All code chunks are set to `eval=FALSE` so they do not execute during bookdown compilation. The processed plotting data is saved to `.Rdata` files which are then loaded in subsequent analysis chapters.

## Load the Data

Load the data objects prepared in previous chapters (MAG catalogue, sample metadata, sequencing statistics, and color schemes).

```{r load_data_02, eval=FALSE}
load("data/MAG_catalogue/data.Rdata")
# load("data/micro/data.Rdata")
load("data/macro/data.Rdata")
load("data/data_colors.Rdata")
```

## Sequencing Statistics Data Frames

This section combines sequencing statistics with sample metadata to create data frames ready for plotting.

### Micro-samples

**WARNING: Micro-scale data not produced for Adenovirus yet. Ignore all code chunks for micro-scale data.**

```{r plot_data_stats_micro_02, warning=FALSE, message=FALSE, eval=FALSE}
# Combine sequencing stats with sample metadata for micro-samples
# NOTE: This is not used for Adenovirus project (micro-scale data not available)
# # Combine sequencing stats with sample metadata
# plot_data_stats <- final_combined_stats %>%
#   left_join(sample_metadata, by = join_by(microsample == microsample))
```

### Macro-samples

```{r plot_data_stats_macro_02, warning=FALSE, message=FALSE, eval=FALSE}
# Combine sequencing statistics with sample metadata for macro-samples
plot_data_stats_macro <- final_combined_stats_macro %>%
  left_join(sample_metadata_macro, by = join_by(microsample == sample)) 

# Set factor levels for proper ordering in plots
plot_data_stats_macro <- plot_data_stats_macro %>%
  mutate(age = factor(age, levels = c(0, 3, 5, 7, 14))) %>%
  mutate(treatment_expl = factor(treatment_expl, levels = c("Control", "Disruption&Pathogen", "Pathogen"))) 
```

## Tidy Tables for Plotting

Convert wide-format data into long (tidy) format for easier plotting with ggplot2. Tidy format has one observation per row, making it ideal for faceting and grouping in visualizations.

### Before-After Trimming Comparison

### Before-after
#### Micro-samples
WARNING: MICRO-SCALE DATA NOT PRODUCED FOR ADENOVIRUS YET, IGNORE ALL CODE CHUNKS FOR MICRO-SCALE DATA

```{r tidy_plot_data_stats_before_after_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# tidy_plot_data_stats_before_after <- plot_data_stats %>%
#   pivot_longer(
#     cols = c(
#     total_sequences_before_trim, total_sequences_after_trim,
#     percent_gc_before_trim, percent_gc_after_trim,
#     total_bases_before_trim, total_bases_after_trim),
#     names_to = c("Metric", "Condition"), # Split column names
#     names_pattern = "(.*)_(before|after)_trim") %>%
#   pivot_wider(names_from = Metric, values_from = value) %>%
#   mutate(Condition = factor(Condition, levels = c("before", "after")))
```

#### Macro-samples
```{r tidy_plot_data_stats_before_after_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
tidy_plot_data_stats_before_after_macro <- plot_data_stats_macro %>%
  pivot_longer(
    cols = c(
    total_sequences_before_trim, total_sequences_after_trim,
    percent_gc_before_trim, percent_gc_after_trim,
    total_bases_before_trim, total_bases_after_trim),
    names_to = c("Metric", "Condition"), # Split column names
    names_pattern = "(.*)_(before|after)_trim") %>%
  pivot_wider(names_from = Metric, values_from = value) %>%
  mutate(Condition = factor(Condition, levels = c("before", "after")))
```



### Samtools mapping to different references
Make tidy table for counts

#### Micro-samples
WARNING: MICRO-SCALE DATA NOT PRODUCED FOR ADENOVIRUS YET, IGNORE ALL CODE CHUNKS FOR MICRO-SCALE DATA

```{r tidy_plot_data_stats_counts_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# category_counts <- tibble::tibble(
#   category = c(
#     "bacteria_total_mapped", "chicken_total_mapped",  
#     "human_total_mapped", "swine_total_mapped",
#     "unmapped", "removed_sequences_after_trim"),
#   label = c("Bacteria", "Chicken", "Human",  "Swine",  "Unmapped", "Removed"),   
#   color = c( "#B92D65", "#E8BD50", "#1092b4", "#3baf6f", "#8d98ae", "#2b2d42"))  
# 
# tidy_plot_data_stats_counts <- plot_data_stats %>%
#   pivot_longer(
#     cols = c(
#       chicken_total_mapped, human_total_mapped,
#       swine_total_mapped, bacteria_total_mapped,
#       unmapped, removed_sequences_after_trim),
#     names_to = "mapping_status", values_to = "counts") %>%
#   left_join(category_counts, by = c("mapping_status" = "category")) %>%
#   # order the groups in desired order
#   mutate(label = factor(label,
#          levels = c("Bacteria", "Chicken", "Human", "Swine", "Unmapped", "Removed")))
# 
# # set labels & colours as colour palette
# color_palette_tidy_plot_data_stats_counts <- setNames(category_counts$color, category_counts$label)
```

#### Macro-samples
```{r tidy_plot_data_stats_counts_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
category_counts <- tibble::tibble(
  category = c(
    "bacteria_total_mapped", "adenovirus_total_mapped", "turkey_total_mapped", "chicken_total_mapped",  
    "human_total_mapped", "swine_total_mapped", 
    "unmapped", "removed_sequences_after_trim"),
  label = c("Bacteria", "Adenovirus", "Turkey","Chicken", "Human",  "Swine",  "Unmapped", "Removed"),   
  color = c( "#B92D65", "#8b4e2e", "#6390fb", "#E8BD50", "#1092b4", "#3baf6f", "#8d98ae", "#2b2d42"))  

tidy_plot_data_stats_counts_macro <- plot_data_stats_macro %>%
  pivot_longer(
    cols = c(
      chicken_total_mapped, human_total_mapped,
      swine_total_mapped, bacteria_total_mapped, turkey_total_mapped, adenovirus_total_mapped,
      unmapped, removed_sequences_after_trim),
    names_to = "mapping_status", values_to = "counts") %>%
  left_join(category_counts, by = c("mapping_status" = "category")) %>%
  # order the groups in desired order
  mutate(label = factor(label,
         levels = c("Bacteria", "Adenovirus", "Turkey", "Chicken", "Human", "Swine", "Unmapped", "Removed")))

# set labels & colours as colour palette
color_palette_tidy_plot_data_stats_counts <- setNames(category_counts$color, category_counts$label)
```


### Percentages of different references

#### Micro-samples
WARNING: MICRO-SCALE DATA NOT PRODUCED FOR ADENOVIRUS YET, IGNORE ALL CODE CHUNKS FOR MICRO-SCALE DATA

```{r tidy_plot_data_stats_counts_prc_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# category_prc <- tibble::tibble(
#   category = c(
#     "bacteria_percentage", "chicken_percentage",  
#     "human_percentage", "swine_percentage",
#     "unmapped_percentage", "trimmed_reads_percentage"),
#   label = c("Bacteria", "Chicken", "Human",  "Swine",  "Unmapped", "Removed"),   
#   color = c( "#B92D65", "#E8BD50", "#1092b4", "#3baf6f", "#8d98ae", "#2b2d42"))  
# 
# 
# tidy_plot_data_stats_counts_prc <- plot_data_stats_combined %>%
#   pivot_longer(
#     cols = c(
#       chicken_percentage, human_percentage,
#       swine_percentage, bacteria_percentage,
#       unmapped_percentage, trimmed_reads_percentage),
#     names_to = "mapping_status", values_to = "counts") %>%
#   left_join(category_prc, by = c("mapping_status" = "category")) %>%
#   # order the groups in desired order
#   mutate(label = factor(label,
#          levels = c("Removed", "Swine", "Human", "Chicken", "Bacteria", "Unmapped")))
# 
# 
# # set labels & colours as colour palette
# color_palette_tidy_plot_data_stats_counts <- setNames(category_counts$color, category_counts$label)
```

##### Macro-samples
```{r tidy_plot_data_stats_counts_prc_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
category_prc <- tibble::tibble(
  category = c(
    "bacteria_percentage", "adenovirus_percentage", "turkey_percentage", "chicken_percentage",  
    "human_percentage", "swine_percentage", 
    "unmapped_percentage", "trimmed_reads_percentage"),
  label = c("Bacteria", "Adenovirus", "Turkey","Chicken", "Human",  "Swine",  "Unmapped", "Removed"),   
  color = c( "#B92D65", "#8b4e2e", "#6390fb", "#E8BD50", "#1092b4", "#3baf6f", "#8d98ae", "#2b2d42"))  


tidy_plot_data_stats_counts_prc_macro <- plot_data_stats_macro %>%
  pivot_longer(
    cols = c(
      chicken_percentage, human_percentage,
      swine_percentage, bacteria_percentage, turkey_percentage, adenovirus_percentage,
      unmapped_percentage, trimmed_reads_percentage),
    names_to = "mapping_status", values_to = "counts") %>%
  left_join(category_prc, by = c("mapping_status" = "category")) %>%
  # order the groups in desired order
  mutate(label = factor(label,
         levels = c("Bacteria", "Adenovirus", "Turkey", "Chicken", "Human", "Swine", "Unmapped", "Removed")))


# set labels & colours as colour palette
color_palette_tidy_plot_data_stats_counts <- setNames(category_counts$color, category_counts$label)
```



## Tidy genome counts

### Micro-samples
WARNING: MICRO-SCALE DATA NOT PRODUCED FOR ADENOVIRUS YET, IGNORE ALL CODE CHUNKS FOR MICRO-SCALE DATA

Make tidy tables for genome counts
```{r tidy_plot_genome_counts_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# tidy_plot_genome_counts <-  genome_counts %>%
#     mutate_at(vars(-genome), ~ . / sum(.)) %>% # scale each sample to 1
#     pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
#     left_join(genome_metadata, by = join_by(genome == genome)) %>% 
#     left_join(sample_metadata, by = join_by(microsample == microsample)) %>% 
#     left_join(final_combined_stats, by = join_by(microsample == microsample)) %>%
#     # Flatten section if it's a list
#     mutate(section = unlist(section)) %>% 
#     # Filter out rows with count <= 0 
#     # (that are redundant after pivot_longer, but important for b div)
#     filter(count > 0) %>%
#     mutate(phylum = factor(phylum, levels = phylum_level_vector),
#            order = factor(order, levels = order_level_vector))
```

```{r tidy_plot_genome_counts_filt_30_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# tidy_plot_genome_counts_filt_30 <-  genome_counts_filt_30 %>%
#     pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
#     left_join(genome_metadata, by = join_by(genome == genome)) %>% 
#     left_join(sample_metadata, by = join_by(microsample == microsample)) %>% 
#     left_join(final_combined_stats, by = join_by(microsample == microsample)) %>%
#     mutate(section = unlist(section)) %>% 
#     filter(count > 0) %>%
#     mutate(phylum = factor(phylum, levels = phylum_level_vector),
#            order = factor(order, levels = order_level_vector))
```

```{r tidy_plot_genome_counts_filt_30_closed_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# tidy_plot_genome_counts_filt_30_closed <-  genome_counts_filt_30 %>%
#     mutate_at(vars(-genome), ~ . / sum(.)) %>% 
#     pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
#     left_join(genome_metadata, by = join_by(genome == genome)) %>% 
#     left_join(sample_metadata, by = join_by(microsample == microsample)) %>% 
#     left_join(final_combined_stats, by = join_by(microsample == microsample)) %>%
#     mutate(section = unlist(section)) %>% 
#     filter(count > 0) %>%
#     mutate(phylum = factor(phylum, levels = phylum_level_vector),
#            order = factor(order, levels = order_level_vector))
```

### Macro-samples

```{r tidy_plot_genome_counts_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
tidy_plot_genome_counts_macro <-  genome_counts_macro %>%
    mutate_at(vars(-genome), ~ . / sum(.)) %>% # scale each sample to 1
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% 
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))
```

```{r tidy_plot_genome_counts_filt_30_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
tidy_plot_genome_counts_filt_30_macro <-  genome_counts_macro_filt_30 %>%
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% # I have changed to the correct metadata
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))
```

```{r tidy_plot_genome_counts_filt_30_macro_closed_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
tidy_plot_genome_counts_filt_30_macro_closed <-  genome_counts_macro_filt_30 %>%
    mutate_at(vars(-genome), ~ . / sum(.)) %>% 
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% # I have changed to the correct metadata
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))
```



## Read count tables (length corrected) dataframes

### Macro-data
Make tidy tables for read counts (length corrected)
```{r tidy_plot_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# unfiltered data
tidy_plot_c_read_counts_macro <-  c_read_counts_macro %>%
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% 
    left_join(final_combined_stats_macro, by = join_by(microsample == microsample)) %>%
    # Filter out rows with count <= 0 
    # (that are redundant after pivot_longer, but important for b div)
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))

# filtered data
tidy_plot_c_read_counts_macro_filt_30 <-  c_read_counts_macro_filt_30 %>%
    pivot_longer(-genome, names_to = "microsample", values_to = "count") %>% 
    left_join(genome_metadata, by = join_by(genome == genome)) %>% 
    left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>% 
    left_join(final_combined_stats_macro, by = join_by(microsample == microsample)) %>%
    # Filter out rows with count <= 0 
    # (that are redundant after pivot_longer, but important for b div)
    filter(count > 0) %>%
    mutate(phylum = factor(phylum, levels = phylum_level_vector),
           order = factor(order, levels = order_level_vector))
```


## Alpha diversity dataframes

Define function to estimate the genomes' alpha diversity measurements from genome counts.
```{r calculate_alpha_diversity_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
calculate_alpha_diversity <- function(input_data, dataset_name) {
  # Step 1: Transform the input data (remove rownames if needed)
  input_data_matrix <- input_data %>%
    column_to_rownames(var = "genome")

  # Step 2: Calculate richness (q = 0)
  richness <- hilldiv(input_data_matrix, q = 0) %>%
    t() %>%
    as.data.frame() %>%
    rename(richness = 1) %>%
    rownames_to_column(var = "microsample")

  # Step 3: Calculate neutral diversity (q = 1)
  neutral <- hilldiv(input_data_matrix, q = 1) %>%
    t() %>%
    as.data.frame() %>%
    rename(neutral = 1) %>%
    rownames_to_column(var = "microsample")

  # Step 4: Calculate phylogenetic diversity (q = 1, with genome tree)
  phylogenetic <- hilldiv(input_data_matrix, q = 1, tree = genome_tree) %>%
    t() %>%
    as.data.frame() %>%
    rename(phylogenetic = 1) %>%
    rownames_to_column(var = "microsample")

  # Step 5: Merge all diversity metrics
  alpha_diversity <- richness %>%
    full_join(neutral, by = "microsample") %>%
    full_join(phylogenetic, by = "microsample") # %>%
    # left_join(sample_metadata, by = "microsample") %>%
    # left_join(final_combined_stats, by = "microsample") %>%

  # Return the alpha_diversity data frame
  return(alpha_diversity)
}
```

Estimate the alpha diversity on the unfiltered and the coverage-filtered counts

### Micro-data
WARNING: MICRO-SCALE DATA NOT PRODUCED FOR ADENOVIRUS YET, IGNORE ALL CODE CHUNKS FOR MICRO-SCALE DATA

```{r alpha_diversities_filtered_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# alpha_div_unfiltered <- calculate_alpha_diversity(
#   input_data = genome_counts,
#   dataset_name = "unfiltered") %>%
#   mutate(filter_level = "unfiltered")
# 
# alpha_div_filtered_30 <- calculate_alpha_diversity(
#   input_data = genome_counts_filt_30,
#   dataset_name = "filtered")  %>% 
#   mutate(filter_level = "filtered_30")
``` 


### Macro-data

```{r alpha_diversities_filtered_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
alpha_div_macro_unfiltered <- calculate_alpha_diversity(
  input_data = genome_counts_macro,
  dataset_name = "unfiltered") %>%
  mutate(filter_level = "unfiltered")

alpha_div_macro_filtered_30 <- calculate_alpha_diversity(
  input_data = genome_counts_macro_filt_30,
  dataset_name = "filtered")  %>% 
  mutate(filter_level = "filtered_30")
```

### Micro-data

```{r alpha_diversity_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# alpha_diversity <- bind_rows(alpha_div_unfiltered, alpha_div_filtered_30) %>%
#   left_join(sample_metadata, by = "microsample") %>%
#   left_join(final_combined_stats, by = "microsample") %>%
#   mutate(filter_level = factor(filter_level, 
#                                levels = c("unfiltered","filtered_30")))
```

Make tidy a-diversity
```{r tidy_plot_alpha_diversity_genomes_micro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# tidy_plot_alpha_diversity_genomes <- alpha_diversity %>%
#   pivot_longer(
#     cols = c(richness, neutral, phylogenetic), 
#     names_to = "metric",                      
#     values_to = "value"                       
#   ) %>%
#       mutate(metric = factor(metric, levels = c("richness", "neutral", "phylogenetic")))
```


### Macro-data

```{r alpha_diversity_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
alpha_diversity_macro <- bind_rows(
  alpha_div_macro_unfiltered,
  alpha_div_macro_filtered_30) %>%
  left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>%
  left_join(final_combined_stats_macro, by = "microsample") %>%
  mutate(filter_level = factor(filter_level, levels = c("unfiltered", 
                                                        "filtered_30")))
```

```{r tidy_plot_alpha_diversity_genomes_macro_02, warning=FALSE, comments="", message=FALSE, eval=FALSE}
tidy_plot_alpha_diversity_genomes_macro <- alpha_diversity_macro %>%
  pivot_longer(
    cols = c(richness, neutral, phylogenetic), 
    names_to = "metric",                      
    values_to = "value"                       
  ) %>%
      mutate(metric = factor(metric, levels = c("richness", "neutral", "phylogenetic")))
```


## Wrap working objects
In the last step, the objects that are needed for downstream analyses are stored in an R object.
```{r wrap_objects_plot_02, eval=FALSE}
## WARNING: MICRO-SCALE DATA NOT PRODUCED FOR ADENOVIRUS YET, IGNORE ALL CODE CHUNKS FOR MICRO-SCALE DATA
# save(
#   plot_data_stats, plot_data_stats_macro, plot_data_stats_combined,
#   tidy_plot_data_stats_before_after, tidy_plot_data_stats_counts, color_palette_tidy_plot_data_stats_counts,
#   # tidy_total_reads,
#   tidy_plot_genome_counts,
#   tidy_plot_genome_counts_filt_30,
#   tidy_plot_genome_counts_filt_30_closed,
#   tidy_plot_alpha_diversity_genomes,
#   file = "data/micro/plot_data.Rdata"
# )

# Save only plot data objects needed for generating visualizations
# These are the processed/tidy data frames used directly in plotting chapters
save(
  sample_metadata_macro,
  plot_data_stats_macro,
  tidy_plot_data_stats_counts_macro,
  tidy_plot_data_stats_counts_prc_macro,
  color_palette_tidy_plot_data_stats_counts,
  tidy_plot_genome_counts_macro,
  tidy_plot_genome_counts_filt_30_macro,
  tidy_plot_genome_counts_filt_30_macro_closed,
  alpha_div_macro_unfiltered,
  alpha_div_macro_filtered_30,
  alpha_diversity_macro,
  tidy_plot_alpha_diversity_genomes_macro,
  tidy_plot_c_read_counts_macro,
  tidy_plot_c_read_counts_macro_filt_30,
  file = "data/macro/plot_data.Rdata"
)

```



