# MAG Catalogue Overview

## Overview

This chapter provides an overview of the Metagenome-Assembled Genome (MAG) catalogue used in this analysis. It displays quality metrics, taxonomic composition, and other properties of the 251 species-representative genomes.

## Load Data

```{r load_data_03, warning=FALSE, message=FALSE}
# Load MAG catalogue data and color schemes
load("data/MAG_catalogue/data.Rdata")
load("data/data_colors.Rdata")
```


## Genome Quality Summary

This section displays summary statistics for genome quality metrics (completeness and contamination) across all MAGs in the catalogue.

```{r genome_quality_summary_03, message=FALSE, warning=FALSE, comment=""}
# Calculate and display summary statistics for genome quality metrics
genome_quality_summary <- data.frame(
  Metric = c("Completeness", "Contamination"),
  Median = c(median(genome_metadata$completeness, na.rm = TRUE), 
             median(genome_metadata$contamination, na.rm = TRUE)),
  Q1 = c(quantile(genome_metadata$completeness, 0.25, na.rm = TRUE),
         quantile(genome_metadata$contamination, 0.25, na.rm = TRUE)),
  Q3 = c(quantile(genome_metadata$completeness, 0.75, na.rm = TRUE),
         quantile(genome_metadata$contamination, 0.75, na.rm = TRUE))
) %>%
  mutate(IQR = Q3 - Q1)

knitr::kable(genome_quality_summary, digits = 2, caption = "Summary statistics (median and IQR) for genome quality metrics")
```


## Genome Circularity and Taxonomic Distribution

```{r genome_circularity_03, message=FALSE, warning=FALSE, comment=""}
# Calculate circularity statistics for all genomes
genome_circularity_stats <- genome_metadata %>%
  summarise(
    circular_count = sum(circularity == "1"),
    total_count = n(),
    circular_percent = 100 * circular_count / total_count
  )

# Display circularity statistics
knitr::kable(genome_circularity_stats, digits = 2, caption = "Genome circularity statistics")

# Calculate taxonomic distribution statistics by order
order_distribution_stats <- order_metadata %>%
  summarise(
    genera_count = sum(number_of_genera),
    lachnospirales_count = sum(number_of_genera[order == "Lachnospirales"]),
    lachnospirales_perc = 100 * lachnospirales_count / genera_count,
    oscillospirales_count = sum(number_of_genera[order == "Oscillospirales"]),
    oscillospirales_perc = 100 * oscillospirales_count / genera_count
  )

# Display taxonomic distribution
knitr::kable(order_distribution_stats, digits = 2, caption = "Taxonomic distribution: Number of genera in major orders")
``` 
  
## Genome Quality Visualization

This section creates visualizations showing the relationship between completeness and contamination, colored by taxonomic order.

```{r genome_quality_plot_order_level_03, message=FALSE, warning=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
# Generate quality biplot showing completeness vs contamination
# Points are colored by taxonomic order and sized by genome length
genome_biplot <- genome_metadata %>%
  select(c(genome,domain,order,completeness,contamination,length)) %>%
  arrange(match(genome, rev(genome_tree$tip.label))) %>% #sort MAGs according to phylogenetic tree
  ggplot(aes(x=completeness,y=contamination,size=length,color=order)) +
              geom_point(alpha=0.7) +
                    xlim(c(70,100)) +
                    ylim(c(25,0)) +
                    scale_color_manual(values=order_colors) +
                    labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                    theme(legend.position = "none")

#Generate contamination boxplot
genome_contamination <- genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(25,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_void() +
                    theme(legend.position = "none",
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(70,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_void() +
                theme(legend.position = "none",
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

# Align margins
genome_biplot <- genome_biplot + theme(plot.margin = margin(5, 5, 5, 5))
genome_completeness <- genome_completeness + theme(plot.margin = margin(0, 5, 0, 5))
genome_contamination <- genome_contamination + theme(plot.margin = margin(5, 0, 5, 0))

# Compose layout
top_row <- genome_completeness + plot_spacer() + plot_layout(widths = c(15, 1))
bottom_row <- genome_biplot + genome_contamination + plot_layout(widths = c(15, 1))

# Stack rows
MAG_plot <- (top_row / bottom_row + plot_layout(heights = c(2, 15))) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom", legend.box = "vertical") 

MAG_plot


```


## MAG catalogue phyla and order colours
```{r phyla_order_colors_plot_03, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.fullwidth=TRUE}
MAG_tax_df <- order_metadata %>%
  select(order, phylum) %>%
  distinct()  # ensure no duplicates

# get the phylum order that facet_grid2 will use (alphabetical by default)
phylum_order <- MAG_tax_df %>%
  distinct(phylum) %>%
  arrange(phylum) %>%
  pull(phylum)

# phylum strip colors in correct order
strip_colors <- phylum_colors[phylum_order]

# plot: use y = 1 directly
ggplot(MAG_tax_df, aes(x = order, y = 1, fill = order)) +
  geom_col(width = 0.9, show.legend = FALSE) +
  geom_text(aes(label = order),
            angle = 90, color = "white", size = 3,
            hjust = 1.1, vjust = 0.5) +
  scale_fill_manual(values = order_colors) +
  facet_grid2(
    . ~ phylum,
    scales = "free_x",
    space = "free",
    strip = strip_themed(
      background_x = elem_list_rect(fill = strip_colors),
      text_x = elem_list_text(angle = 90, color = "white", face = "bold")
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    panel.spacing = unit(0.3, "lines")
  )
```

## MAG catalogue phyla and order colours, and genera number
```{r phyla_order_genera_plot_03, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.fullwidth=TRUE}
MAG_tax_df <- order_metadata %>%
  select(phylum, order, number_of_genera) %>%
  distinct() %>%
  mutate(
    order_color = order_colors[order],
    phylum_color = phylum_colors[phylum]
  )

# 2. Get facet strip colors in correct order
phylum_order <- MAG_tax_df %>%
  distinct(phylum) %>%
  arrange(phylum) %>%
  pull(phylum)

strip_colors <- unname(phylum_colors[phylum_order])

# 3. Plot
ggplot(MAG_tax_df, aes(x = order, y = number_of_genera, fill = order)) +
  # Bars
  geom_col(width = 0.9, show.legend = FALSE) +
  
  # Color scale
  scale_fill_manual(values = order_colors) +
  
  # Facet by phylum (strip on bottom)
  facet_grid2(
    . ~ phylum,
    scales = "free_x",
    space = "free_x",
    switch = "x", 
    strip = strip_themed(
      background_x = elem_list_rect(fill = strip_colors),
      text_x = elem_list_text(angle = 90, color = "white", face = "bold")
    )
  ) +

  # Flip x-axis title and text to top
  scale_x_discrete(position = "top") +

  # Theme
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
    axis.title = element_blank(),
    panel.spacing = unit(1, "lines"),
    strip.background = element_blank(),
    strip.placement = "outside"
  )
```


## Genome phylogeny plot
```{r genome_phylogeny_plot_03, message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
# Generate the phylum color heatmap
phylum_heatmap <- genome_metadata %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")

# Generate the order color heatmap
order_heatmap <- genome_metadata %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome, order) %>%
    mutate(order = factor(order, levels = unique(order))) %>%
    column_to_rownames(var = "genome")

# Generate the basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.1)

# Add the phylum ring
circular_tree <- gheatmap(circular_tree, 
                          phylum_heatmap, 
                          offset = 1.3, width = 0.2, colnames = FALSE) +
    scale_fill_manual(values = phylum_colors, name = "Phylum") +
    geom_tiplab2(size = 1, hjust = -0.1) +
    theme(plot.margin = margin(0, 0, 0, 0), 
          panel.margin = margin(0, 0, 0, 0))

# Flush color scale for the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add order ring
circular_tree <- gheatmap(circular_tree, 
                          order_heatmap, 
                          offset = 0.55, width = 0.3, colnames = FALSE) +
    scale_fill_manual(values = order_colors, name = "Order") +
    theme(legend.position = "right")

# # Flush color scale to enable a new color scheme in the next ring
# circular_tree <- circular_tree + new_scale_fill()
# 
# circular_tree <- circular_tree +
#     new_scale_fill() +
#     scale_fill_manual(values = c("1" = "#fcbb6d", "0" = "#465c7a"), 
#                       name = 'Circularized genomes') +
#     geom_fruit(
#         data = genome_metadata,
#         geom = geom_bar,
#         # Use as.factor to map circularity
#         mapping = aes(x = 1, y = genome, fill = circularity),  
#         offset = 1.0,
#         orientation = "y",
#         stat = "identity")


# Add text
circular_tree <-  circular_tree +
  annotate('text', x=3.0, y=0, label='           Order', size=3.5) +
  annotate('text', x=3.4, y=0, label='              Phylum', size=3.5) #+
  # annotate('text', x=4.2, y=0, label='                      Circularised', size=3.5)

#Plot circular tree
p <-circular_tree %>% open_tree(45) %>% rotate_tree(90) +
  theme(legend.position = "bottom", legend.box = "vertical")
# print(p)
# ggsave(filename = "results/figures/MAG_macro_tree.pdf", plot = p,
#       units = "cm", width = 29, height = 40)

```
## Genome phylogeny plot without legend
```{r genome_phylogeny_no_legend_03, message=FALSE, warning=FALSE, fig.height=7, fig.width=7, fig.fullwidth=TRUE}
p +
  plot_layout(guides = "collect") &
  theme(legend.position = "none")
```



```{r function_ordination_gifts_aggregate_03, message=FALSE, fig.height=14, fig.width=10}

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts,GIFT_db)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

```


## Functional overview

```{r function_group_colors_03, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}

function_group_colors <- c(
  B01 = "#f7941d",  # topmost color (orange)
  B02 = "#f26c1f",  # apricot-orange
  B03 = "#e63e26",  # red-orange
  B04 = "#e6004c",  # deep pink
  B06 = "#a1246b",  # magenta
  B07 = "#ed559f",  # pink
  B08 = "#f99388",  # peach
  B09 = "#1a3c85",  # dark blue
  B10 = "#2b8ac4",  # blue

  D01 = "#5ec4ed",  # sky blue
  D02 = "#89d2ef",  # pale blue
  D03 = "#d9efe9",  # pale mint
  D04 = "#9ed7b4",  # light green
  D05 = "#4fb18f",  # mint green
  D06 = "#3b946d",  # dark teal
  D07 = "#2a7854",  # deeper teal
  D08 = "#295b47",  # forest green
  D09 = "#234c3f",  # deep green

  S01 = "#1b3e34",  # nearly black-green
  S02 = "#aaaaaa",  # light grey
  S03 = "#666666"   # dark grey
)
```

```{r function_heatmap_tree_03, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
order_heatmap <- genome_metadata %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,order) %>%
    column_to_rownames(var = "genome")

# Aggregate basal GIFT into elements
function_table <- genome_gifts %>%
    to.elements(., GIFT_db)

# 1. Base tree
function_tree <- force.ultrametric(genome_tree, method="extend") %>%
                ggtree(size = 0.3) 

# 2. First heatmap: add order (row) color strip
function_tree <- gheatmap(
    function_tree, 
    order_heatmap, 
    offset = 0, width = 0.1, colnames = FALSE
  ) +
  scale_fill_manual(values = order_colors) +
  labs(fill = "Order")

# 5. Reset fill scale again for main function matrix
function_tree <- function_tree + new_scale_fill()

# 6. Add the full function table heatmap (numeric)
function_tree <- gheatmap(
    function_tree, 
    function_table, 
    offset = 0.5, 
    width = 3.5, 
    colnames = FALSE
  ) +
  vexpand(0) + #   vexpand(0.08) +
  coord_cartesian(clip = "off") +
  scale_fill_gradient(low = "#f4f4f4", high = "#666666", na.value = "white") +
  labs(fill = "GIFT")

# 7. Final plot
function_tree <- function_tree +
  theme_void() +
  theme(legend.position = "none") 
# function_tree
```



```{r function_heatmap_annotation_03, message=FALSE, warning=FALSE, fig.height=2, fig.width=10, fig.fullwidth=TRUE}
# Join to get Code_function (e.g. B01, D01) for each Code_element (e.g. B0101, B0204)
column_annotation <- tibble(Code_element = colnames(function_table)) %>%
  left_join(
    GIFT_db %>% 
      select(Code_element, Code_function) %>% 
      distinct(Code_element, .keep_all = TRUE),
    by = "Code_element"
  )

# Create a 1-row matrix with Code_function values
column_annotation_matrix <- matrix(
  column_annotation$Code_function,
  nrow = 1,
  dimnames = list("Function_group", column_annotation$Code_element)
)
# Ensure Code_element order matches your heatmap
column_annotation_df <- column_annotation %>%
  distinct(Code_element, Code_function)

column_annotation_df$Code_element <- factor(
  column_annotation_df$Code_element,
  levels = column_annotation$Code_element
)
# Plot annotation strip
annotation_strip <- ggplot(column_annotation_df, aes(x = Code_element, y = 1, fill = Code_function)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = function_group_colors, na.translate = FALSE) +
  theme_void() +
  theme(
    axis.text.x = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  ) +
  guides(fill = "none") +
  theme_void()


# annotation_strip

```

```{r function_heatmap_combined_03, message=FALSE, warning=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
combined_plot <- ((plot_spacer() | wrap_elements(annotation_strip) | plot_spacer()) + plot_layout(widths = c(10.1, 25, 1.1))) /
                 wrap_elements(function_tree) +
                 plot_layout(heights = c(0.2, 10))  # adjust strip height as needed

combined_plot

```

## Functional distances

```{r function_distances, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
functional_distances <- genome_gifts %>%
    to.elements(., GIFT_db) %>%
    as.data.frame() %>%
    stats::dist(., method = "manhattan") / ncol(genome_gifts[genome_metadata$genome, ])

mean(functional_distances)
sd(functional_distances)
max(functional_distances)
```


## Functional ordination 

PCoA functional ordination with PCA loadings.

```{r function_ordination_pcoa, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
gift_pcoa <- genome_gifts %>%
    to.elements(., GIFT_db) %>%
    as.data.frame() %>%
    vegdist(method="euclidean") %>%
    pcoa()

gift_pcoa_rel_eigen <- gift_pcoa$values$Relative_eig[1:10]


# Get genome positions
gift_pcoa_vectors <- gift_pcoa$vectors %>% #extract vectors
  as.data.frame() %>% 
  select(Axis.1,Axis.2) # keep the first 2 axes

gift_pcoa_eigenvalues <- gift_pcoa$values$Eigenvalues[c(1,2)]

gift_pcoa_gifts <- cov(genome_gifts, scale(gift_pcoa_vectors)) %*% diag((gift_pcoa_eigenvalues/(nrow(genome_gifts)-1))^(-0.5)) %>%
  as.data.frame() %>% 
  rename(Axis.1=1,Axis.2=2) %>% 
  rownames_to_column(var="label") %>% 
  #get function summary vectors
  mutate(func=substr(label,1,3)) %>% 
  group_by(func) %>% 
  summarise(Axis.1=mean(Axis.1),
            Axis.2=mean(Axis.2)) %>% 
  rename(label=func) %>% 
  filter(!label %in% c("S01","S02","S03"))

```


```{r function_ordination_pcoa_plot, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}

scale <- 15 # scale for vector loadings
pcoa_plot <- gift_pcoa_vectors %>% 
  rownames_to_column(var="genome") %>% 
  left_join(genome_metadata, by="genome") %>%
  ggplot() +
      #genome positions
      scale_color_manual(values=order_colors)+
      geom_point(aes(x=Axis.1,y=Axis.2, color=order, size=length), 
                 alpha=0.9, shape=16) +
    guides(color = "none") +
      #scale_color_manual(values=phylum_colors) +
      scale_size_continuous(range = c(0.1,5)) +
      #loading positions
      geom_segment(data=gift_pcoa_gifts, 
                   aes(x=0, y=0, xend=Axis.1 * scale, yend=Axis.2 * scale),
                    arrow = arrow(length = unit(0.3, "cm"), 
                    type = "open", 
                    angle = 25),
                    linewidth = 0.5, 
                    color = "black") +
     #Primary and secondary scale adjustments
     # scale_x_continuous(name = paste0("PCoA1 (",round(gift_pcoa_rel_eigen[1]*100, digits = 2), " %)"),
     #                  sec.axis = sec_axis(~ . / scale)
     #        ) +
     # scale_y_continuous(name = paste0("PCoA2 (",round(gift_pcoa_rel_eigen[2]*100, digits = 2), " %)"),
     #                  sec.axis = sec_axis(~ . / scale)
     #        ) +
    geom_label_repel(data = gift_pcoa_gifts,
                     aes(label = label, x = Axis.1 * scale, y = Axis.2 * scale),
                     segment.color = 'transparent') +
    theme_minimal() 


# pcoa_plot

```

```{r function_ordination_pcoa_length_curve, message=FALSE, fig.height=2, fig.width=10, fig.fullwidth=TRUE}
gift_axis_1 <- gift_pcoa_vectors %>% 
  rownames_to_column(var="genome") %>% 
  left_join(genome_metadata, by="genome") %>%
  ggplot(aes(x=Axis.1, y=length)) +
      geom_smooth(color = "black") +
      theme_minimal() + 
      theme(legend.position = "none") +
  labs(y = "Genome length", x =  "PCoA Axis 1") 

# gift_axis_1
```


```{r function_ordination_pcoa_D01_plot_03, message=FALSE, fig.height=7, fig.width=2, fig.fullwidth=TRUE}
gift_df_PCoA <- genome_gifts %>%
    to.elements(., GIFT_db) %>% 
    to.functions(., GIFT_db) %>% 
    as.data.frame() %>% 
    rownames_to_column(var="genome") %>% 
    left_join(genome_metadata, by="genome") %>%
    left_join(gift_pcoa_vectors %>% rownames_to_column(var="genome"), by="genome")

gift_long <- gift_df_PCoA %>%
  select(genome, Axis.2,
         B01, B02, B03, B04, B06, B07, B08, B09, B10,
         D01, D02, D03, D04, D05, D06, D07, D08, D09,
         S01, S02, S03) %>%
  pivot_longer(
    cols = c(B01, B02, B03, B04, B06, B07, B08, B09, B10,
             D01, D02, D03, D04, D05, D06, D07, D08, D09,
             S01, S02, S03),
    names_to = "component",
    values_to = "value"
  )
# Plot: x = Axis.2, y = value, colored by component (D01 vs D02)
gift_plot_PC2 <- gift_long %>%
  filter(component %in% c("D01", "D09")) %>%
  ggplot(aes(y = value, x =  Axis.2, color = component)) +
  geom_smooth() +
  scale_color_manual(values = c(D01 = "#5ec4ed", D09 = "#234c3f")) +
  theme_minimal() +
  labs(y = "GIFT", x =  "PCoA Axis 2", color = "Component") +
  theme(legend.position = "right")+ 
  coord_flip()
# gift_plot_PC2
```

```{r function_ordination_pcoa_combined_03, message=FALSE, fig.height=8, fig.width=13, fig.fullwidth=TRUE}
((gift_axis_1| plot_spacer()) + plot_layout(widths = c(4.5, 1))) / 
  ((pcoa_plot | gift_plot_PC2) + plot_layout(widths = c(7, 1))) +
  plot_layout(heights = c(1, 6), guides = "collect") +
  theme(legend.position = "bottom") 

```


```{r GIFT_functions_03, message=FALSE, fig.height=7, fig.width=2, fig.fullwidth=TRUE}
GIFT_db_functions <- GIFT_db %>%
  select(Code_function, Function, Domain) %>%
  distinct(Code_function, .keep_all = TRUE)
GIFT_db_functions
```


```{r GIFT_elements_03, message=FALSE, fig.height=7, fig.width=2, fig.fullwidth=TRUE}
GIFT_db_elements <- GIFT_db %>%
  select(Code_element, Element, Function, Domain) %>%
  distinct(Code_element, .keep_all = TRUE)
GIFT_db_elements

GIFT_db_elements %>% filter(Element == "Butyrate")


# I am searching the species in genome metadata, and the compound from GIFT_db_elements in the GIFTs_elements
# Faecalibacterium intestinigallinarum
GIFTs_elements["D301020:bin_000009", "B0402"] 

# Mediterraneibacter faecipullorum
GIFTs_elements["D301031:bin_000019", "B0402"] 

# Blautia pullicola
GIFTs_elements["D300974:bin_000001", "B0402"] 

# Mediterraneibacter excrementigallinarum
GIFTs_elements["D301051:bin_000001", "B0402"] 

# Coproplasma stercorigallinarum
GIFTs_elements["D300981:bin_000003", "B0402"] 


```

```{r GIFT_legend_03, message=FALSE, fig.height=8, fig.width=3, fig.fullwidth=TRUE}


GIFT_legend_df <- GIFT_db_functions %>%
  mutate(color = function_group_colors[Code_function]) %>%
  arrange(Code_function)  # keep the order consistent

function_colors <- ggplot(GIFT_legend_df, aes(y = reorder(Code_function, desc(Code_function)), x = 1, fill = Code_function)) +
  geom_tile(color = "white", linewidth = 0.4, width = 1.1, height = 0.9) +
  scale_fill_manual(values = function_group_colors, guide = "none") +
  geom_text(aes(label = paste0(Code_function, ": ", Function)), 
            hjust = 0, nudge_x = - 0.5, color = "white", size = 3.5, fontface = "bold") +
  theme_void() +
  theme(
    aspect.ratio = NULL,
    plot.margin = margin(0, 0, 0, 0)
  ) 
function_colors
```


```{r function_ordination_pcoa_components_03, message=FALSE, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
gift_long %>%
  ggplot(aes(x = Axis.2, y = value)) +
  geom_smooth(se = FALSE, color = "black", linewidth = 0.6) +
  facet_wrap(~ component, ncol = 5, scales = "free_y") +
  theme_minimal() +
  labs(x = "PCoA Axis 2", y = "Component value") +
  theme(
    strip.text = element_text(size = 8, face = "bold"),
    panel.spacing = unit(0.5, "lines"),
    axis.title = element_text(size = 10)
  )

```
