# Adenovirus Presence Analysis

## Load Data and Plotting Utilities

```{r load_data_04A, warning=FALSE, message=FALSE}
# Load processed plotting data and color schemes
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
```

## Adenovirus Percentage Over Time

This section visualizes the percentage of sequencing reads mapping to adenovirus genomes across different time points (days post infection) for both broiler and layer breeds.

### Broilers - Adenovirus Percentage

```{r adenovirus_percentage_broiler_stats, warning=FALSE, message=FALSE, comment=""}
# Calculate summary statistics (median, IQR) for adenovirus percentage by treatment and age
var <- "adenovirus_percentage"
summary_stats_broiler <- plot_data_stats_macro %>%
  filter(treatment != "TF0") %>%  # Exclude negative controls
  filter(animal_breed == "broiler") %>%
  group_by(treatment_expl, age) %>%
  summarise(
    median = median(.data[[var]], na.rm = TRUE),
    IQR    = IQR(.data[[var]], na.rm = TRUE),
    Q1     = quantile(.data[[var]], 0.25, na.rm = TRUE),
    Q3     = quantile(.data[[var]], 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label_txt = formatC(median, format = "f", digits = 5))

# Display summary statistics table with proper formatting
knitr::kable(summary_stats_broiler, digits = 5, caption = "Summary statistics (median and IQR) for adenovirus percentage in broilers by treatment and age")
```

```{r adenovirus_percentage_broiler_plot, warning=FALSE, message=FALSE, fig.height=5, fig.width=10}
# Create boxplot with jittered points showing adenovirus percentage over time for broilers
boxplot_reads_prc_1_broiler <- plot_data_stats_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "broiler") %>%
  ggplot(aes(x = age, y = adenovirus_percentage, color = treatment)) +
  geom_boxplot(outlier.shape = NA, color = "black") +  # Boxplot without outliers (shown as jitter)
  geom_jitter(width = 0.05, alpha = 0.5, size = 2) +  # Individual data points
  scale_color_manual(values = treatment_colours_bright) +
  facet_nested(. ~ treatment_expl, scales = "fixed", space = "fixed", switch = "y") +
  labs(x = "Days post infection", y = "% reads mapping to Adenovirus") +
  guides(color = "none") +  # Remove legend
  theme_minimal() +
  custom_ggplot_theme +
  theme(panel.spacing = unit(0.3, "lines"),
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) + 
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  # Add median values as text labels above boxes
  geom_text(
    data = summary_stats_broiler,
    aes(x = age, y = median, label = label_txt),
    inherit.aes = FALSE,
    vjust = -3.0,
    size = 3, alpha = 1, color = "black"
  ) +
  ggtitle("Broilers")

# Display the plot (commented out - used in combined plot)
# boxplot_reads_prc_1_broiler
```

### Layers - Adenovirus Percentage

```{r adenovirus_percentage_layer_stats, warning=FALSE, message=FALSE, comment=""}
# Calculate summary statistics (median, IQR) for adenovirus percentage by treatment and age
var <- "adenovirus_percentage"
summary_stats_layer <- plot_data_stats_macro %>%
  filter(treatment != "TF0") %>%  # Exclude negative controls
  filter(animal_breed == "layer") %>%
  group_by(treatment_expl, age) %>%
  summarise(
    median = median(.data[[var]], na.rm = TRUE),
    IQR    = IQR(.data[[var]], na.rm = TRUE),
    Q1     = quantile(.data[[var]], 0.25, na.rm = TRUE),
    Q3     = quantile(.data[[var]], 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label_txt = formatC(median, format = "f", digits = 5))

# Display summary statistics table with proper formatting
knitr::kable(summary_stats_layer, digits = 5, caption = "Summary statistics (median and IQR) for adenovirus percentage in layers by treatment and age")
```
```{r adenovirus_percentage_layer_plot, warning=FALSE, message=FALSE, fig.height=5, fig.width=10}
# Create boxplot with jittered points showing adenovirus percentage over time for layers
boxplot_reads_prc_1_layer <- plot_data_stats_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "layer") %>%
  ggplot(aes(x = age, y = adenovirus_percentage, color = treatment)) +
  geom_boxplot(outlier.shape = NA, color = "black") +  # Boxplot without outliers (shown as jitter)
  geom_jitter(width = 0.05, alpha = 0.8, size = 2) +  # Individual data points
  scale_color_manual(values = treatment_colours_bright) +
  facet_nested(. ~ treatment_expl, scales = "fixed", space = "fixed", switch = "y") +
  labs(x = "Days post infection", y = "% reads mapping to Adenovirus") +
  guides(color = "none") +  # Remove legend
  theme_minimal() +
  custom_ggplot_theme +
  theme(panel.spacing = unit(0.3, "lines"),
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) + 
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  # Add median values as text labels above boxes
  geom_text(
    data = summary_stats_layer,
    aes(x = age, y = median, label = label_txt),
    inherit.aes = FALSE,
    vjust = -3.0,
    size = 3, alpha = 1, color = "black"
  ) +
  ggtitle("Layers")

# Display the plot (commented out - used in combined plot)
# boxplot_reads_prc_1_layer
```

## Adenovirus Read Counts by Sample

This section shows absolute read counts mapping to adenovirus genomes for each individual sample, grouped by breed and treatment.

### Layers - Adenovirus Read Counts

```{r adenovirus_reads_layer_plot, warning=FALSE, message=FALSE, fig.height=5, fig.width=10}
# Create barplot showing absolute adenovirus read counts per sample for layers
barplot_layer <- plot_data_stats_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "layer") %>%
  ggplot(aes(x = microsample, y = adenovirus_total_mapped, fill = treatment)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = treatment_colours_bright) +
  labs(x = "Samples", y = "Adenovirus reads", fill = "Treatment") +
  facet_nested(. ~ treatment_expl + age, scales = "free", space = "free") +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.x = element_blank(),  # Hide sample IDs for clarity
        axis.text.y = element_text(),
        legend.position = "bottom",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Layers")

# Display the plot (commented out - used in combined plot)
# barplot_layer
```

### Broilers - Adenovirus Read Counts

```{r adenovirus_reads_broiler_plot, warning=FALSE, message=FALSE, fig.height=5, fig.width=10}
# Create barplot showing absolute adenovirus read counts per sample for broilers
barplot_broiler <- plot_data_stats_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "broiler") %>%
  ggplot(aes(x = microsample, y = adenovirus_total_mapped, fill = treatment)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = treatment_colours_bright) +
  labs(x = "Samples", y = "Adenovirus reads", fill = "Treatment") +
  facet_nested(. ~ treatment_expl + age, scales = "free", space = "free") +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.x = element_blank(),  # Hide sample IDs for clarity
        axis.text.y = element_text(),
        legend.position = "bottom",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers")

# Display the plot (commented out - used in combined plot)
# barplot_broiler
```

### Combined Visualization

```{r adenovirus_combined_plot, warning=FALSE, message=FALSE, fig.height=12, fig.width=10}
# Combine all four plots (percentage and read counts for both breeds) in a single figure
# Using patchwork to arrange plots vertically with appropriate heights
(boxplot_reads_prc_1_layer + barplot_layer + 
 boxplot_reads_prc_1_broiler + barplot_broiler) +
  plot_layout(
    ncol = 1,        # Single column layout
    heights = c(2, 1, 2, 1),   # Relative heights: percentage plots taller than barplots
    guides = "collect"  # Collect legends into one
  ) + 
  plot_annotation(tag_levels = 'a') &  # Add panel labels (a, b, c, d)
  theme(legend.position = "bottom", legend.box = "vertical")
```
