# Alpha Diversity Analysis

## Load Data and Plotting Utilities

```{r load_data_04D, warning=FALSE, message=FALSE}
# Load processed plotting data and color schemes
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
```

## Alpha Diversity by Breed

Alpha diversity measures within-sample microbial community diversity. This section compares richness and Shannon diversity metrics between breeds and treatments over time.

### Layers - Alpha Diversity

```{r alpha_diversity_layer_stats, warning=FALSE, message=FALSE, comment=""}
# Calculate summary statistics for alpha diversity metrics by treatment, age, and metric
var <- "value"
summary_stats_layer <- tidy_plot_alpha_diversity_genomes_macro %>%
  filter(animal_breed == "layer") %>%
  filter(filter_level == 'filtered_30') %>%
  filter(filter_status == "Retained by filtering") %>%  # Remove empty samples
  group_by(treatment_expl, age, metric) %>%
  summarise(
    median = median(.data[[var]], na.rm = TRUE),
    IQR    = IQR(.data[[var]], na.rm = TRUE),
    Q1     = quantile(.data[[var]], 0.25, na.rm = TRUE),
    Q3     = quantile(.data[[var]], 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label_txt = formatC(median, format = "f", digits = 1))

# Display summary statistics table
knitr::kable(summary_stats_layer, digits = 3, caption = "Summary statistics (median and IQR) for alpha diversity metrics in layers")
```

```{r alpha_diversity_layer_plot, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Create boxplot showing alpha diversity metrics over time for layers
plot_alpha_div_layer <- tidy_plot_alpha_diversity_genomes_macro %>%
  filter(animal_breed == "layer") %>%
  filter(filter_level == 'filtered_30') %>%
  filter(filter_status == "Retained by filtering") %>%  # Remove empty samples
  ggplot(aes(x = age, y = value, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  scale_fill_manual(
    name = "Treatment",
    breaks = names(treatment_colours_bright),
    values = treatment_colours_bright
  ) +
  geom_jitter(width = 0.1, alpha = 0.6) +
  facet_nested(metric ~ treatment_expl, scales = "free", space = "fixed") +
  labs(x = "Days Post Infection", y = "Alpha diversity value") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  # Add median values as text labels
  geom_text(
    data = summary_stats_layer,
    aes(x = age, y = median, label = label_txt),
    inherit.aes = FALSE,
    vjust = -3.0,
    size = 3, alpha = 1, color = "black"
  ) +
  expand_limits(y = 0) +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Layers: Alpha diversity - 30% coverage filter")

# Display the plot (commented out - used in combined plot)
# plot_alpha_div_layer
```

### Broilers - Alpha Diversity

```{r alpha_diversity_broiler_stats, warning=FALSE, message=FALSE, comment=""}
# Calculate summary statistics for alpha diversity metrics by treatment, age, and metric
var <- "value"
summary_stats_broiler <- tidy_plot_alpha_diversity_genomes_macro %>%
  filter(animal_breed == "broiler") %>%
  filter(filter_level == 'filtered_30') %>%
  filter(filter_status == "Retained by filtering") %>%  # Remove empty samples
  group_by(treatment_expl, age, metric) %>%
  summarise(
    median = median(.data[[var]], na.rm = TRUE),
    IQR    = IQR(.data[[var]], na.rm = TRUE),
    Q1     = quantile(.data[[var]], 0.25, na.rm = TRUE),
    Q3     = quantile(.data[[var]], 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label_txt = formatC(median, format = "f", digits = 1))

# Display summary statistics table
knitr::kable(summary_stats_broiler, digits = 3, caption = "Summary statistics (median and IQR) for alpha diversity metrics in broilers")
```

```{r alpha_diversity_broiler_plot, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Create boxplot showing alpha diversity metrics over time for broilers
plot_alpha_div_broiler <- tidy_plot_alpha_diversity_genomes_macro %>%
  filter(animal_breed == "broiler") %>%
  filter(filter_level == 'filtered_30') %>%
  filter(filter_status == "Retained by filtering") %>%  # Remove empty samples
  ggplot(aes(x = age, y = value, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  scale_fill_manual(
    name = "Treatment",
    breaks = names(treatment_colours_bright),
    values = treatment_colours_bright
  ) +
  geom_jitter(width = 0.1, alpha = 0.6) +
  facet_nested(metric ~ treatment_expl, scales = "free", space = "fixed") +
  labs(x = "Days Post Infection", y = "Alpha diversity value") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  # Add median values as text labels
  geom_text(
    data = summary_stats_broiler,
    aes(x = age, y = median, label = label_txt),
    inherit.aes = FALSE,
    vjust = -3.0,
    size = 3, alpha = 1, color = "black"
  ) +
  expand_limits(y = 0) +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers: Alpha diversity - 30% coverage filter")

# Display the plot (commented out - used in combined plot)
# plot_alpha_div_broiler
```

### Combined Visualization

```{r alpha_diversity_combined, warning=FALSE, message=FALSE, fig.height=10, fig.width=12}
# Adjust theme for side-by-side display (remove duplicate y-axis labels)
plot_alpha_div_layer <- plot_alpha_div_layer +
  theme(
    strip.background.y = element_blank(),
    strip.text.y.right = element_blank()
  )

# Combine both plots side-by-side
(plot_alpha_div_layer | plot_alpha_div_broiler) + 
  plot_annotation(tag_levels = 'a') &
  theme(legend.position = "bottom", legend.box = "vertical")
```

## Detailed Alpha Diversity - Layers with Animal Labels

```{r alpha_diversity_layer_detailed, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Create detailed plot showing individual animals labeled for layers
# This plot uses the neutral (Shannon) diversity metric
tidy_plot_alpha_diversity_genomes_macro %>%
  filter(animal_breed == "layer") %>%
  filter(filter_level == 'filtered_30') %>%
  filter(filter_status == "Retained by filtering") %>%  # Remove empty samples
  filter(metric == "neutral") %>%
  ggplot(aes(x = age, y = value, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  scale_fill_manual(
    name = "Treatment",
    breaks = names(treatment_colours_bright),
    values = treatment_colours_bright
  ) +
  geom_jitter(width = 0.3, alpha = 0.6) +
  facet_nested(treatment_expl ~ ., scales = "free", space = "fixed") +
  # Add animal labels to individual points
  geom_text(
    aes(label = animal), 
    position = position_jitter(width = 0.2, height = 0), 
    size = 3.5, alpha = 0.7, check_overlap = FALSE
  ) +
  expand_limits(y = 0) +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Alpha diversity - neutral - 30% coverage filtering level - LAYERS")

# Display the plot
```


## Relationship Between Genome Length and Alpha Diversity

This section explores the relationship between median genome length per sample and alpha diversity (richness).

### Prepare Data: Calculate Median Genome Length

```{r alpha_genome_length_prep, warning=FALSE, message=FALSE, comment=""}
# Calculate median genome length per sample
# First, remove zero-count entries to get accurate length calculations
tidy_plot_genome_counts_filt_30_macro_closed_zerosrem <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(count > 0)

# Calculate median genome length for each sample
median_lengths <- tidy_plot_genome_counts_filt_30_macro_closed_zerosrem %>%
  group_by(microsample) %>%
  summarise(median_genome_length = median(length, na.rm = TRUE), .groups = "drop")

# Join median lengths with alpha diversity data
alpha_div_macro_filtered_30_gen_length <- alpha_div_macro_filtered_30 %>%
  left_join(median_lengths, by = join_by(microsample == microsample)) %>%
  left_join(sample_metadata_macro, by = join_by(microsample == sample))
```

### Visualize Relationship: Linear Model

```{r alpha_genome_length_scatter, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
# Create scatter plot showing relationship between median genome length and richness
# Points colored by treatment, shaped by days post-infection
alpha_div_macro_filtered_30_gen_length %>%
  filter(treatment != "TF0") %>%  # Exclude negative controls
  ggplot(aes(x = median_genome_length, y = richness, color = treatment, shape = factor(age))) +
  geom_point(size = 3, alpha = 0.8) +
  # Add linear regression line with confidence interval
  geom_smooth(method = "lm", se = TRUE, aes(group = 1), color = "black", linewidth = 1) +
  scale_color_manual(values = treatment_colours_bright, name = "Treatment") +
  scale_shape_manual(
    name = "DPI",
    values = c("0" = 16, "3" = 17, "5" = 15, "7" = 18, "14" = 4)
  ) +
  labs(x = "Median genome length (bp) per sample", y = "Richness") +
  ylim(c(0, 150)) +
  theme_minimal() +
  custom_ggplot_theme

# Display the plot
```

### Statistical Modeling: Negative Binomial Regression

```{r alpha_genome_length_prep_model, warning=FALSE, message=FALSE, comment=""}
# Recalculate median lengths (if needed) and prepare data for modeling
tidy_plot_genome_counts_filt_30_macro_closed_zerosrem <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(count > 0)

median_lengths <- tidy_plot_genome_counts_filt_30_macro_closed_zerosrem %>%
  group_by(microsample) %>%
  summarise(median_genome_length = median(length, na.rm = TRUE), .groups = "drop")

# Join with alpha diversity and filter out controls
alpha_div_macro_filtered_30_gen_length <- alpha_div_macro_filtered_30 %>%
  left_join(median_lengths, by = join_by(microsample == microsample)) %>%
  left_join(sample_metadata_macro, by = join_by(microsample == sample)) %>%
  filter(treatment != "TF0")
```

```{r alpha_genome_length_model_plot, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}

# Fit negative binomial regression model (appropriate for count data)
model_nb <- glm.nb(richness ~ median_genome_length, 
                   data = alpha_div_macro_filtered_30_gen_length)

# Fit null model (intercept only) for comparison
model_null <- glm.nb(richness ~ 1,
                     data = alpha_div_macro_filtered_30_gen_length)

# Calculate McFadden pseudo-R² (model fit metric)
pseudo_r2 <- 1 - (logLik(model_nb) / logLik(model_null))
cat("McFadden pseudo-R² =", round(pseudo_r2, 3), "\n")

# Calculate Spearman correlation (non-parametric)
cor_test <- cor.test(
  ~ richness + median_genome_length,
  data = alpha_div_macro_filtered_30_gen_length,
  method = "spearman"
)

rho  <- cor_test$estimate
pval <- cor_test$p.value
cat("Spearman correlation (ρ) =", round(rho, 3), ", p =", signif(pval, 3), "\n")

# Create prediction data frame for model visualization
newdata <- data.frame(
  median_genome_length = seq(
    min(alpha_div_macro_filtered_30_gen_length$median_genome_length, na.rm = TRUE),
    max(alpha_div_macro_filtered_30_gen_length$median_genome_length, na.rm = TRUE),
    length.out = 200
  )
)

# Get predictions and standard errors on link scale
pred <- predict(model_nb, newdata, type = "link", se.fit = TRUE)

# Back-transform predictions to response scale and calculate 95% confidence intervals
newdata$fit <- exp(pred$fit)
newdata$lwr <- exp(pred$fit - 1.96 * pred$se.fit)   # Lower 95% CI
newdata$upr <- exp(pred$fit + 1.96 * pred$se.fit)   # Upper 95% CI

# Position for statistics annotations
x_pos <- max(alpha_div_macro_filtered_30_gen_length$median_genome_length, na.rm = TRUE) * 0.98
y_pos <- max(alpha_div_macro_filtered_30_gen_length$richness, na.rm = TRUE) * 0.95

# Create plot with fitted negative binomial curve
ggplot(alpha_div_macro_filtered_30_gen_length, 
       aes(x = median_genome_length, y = richness)) +
  # Confidence interval ribbon
  geom_ribbon(data = newdata, 
              aes(x = median_genome_length, ymin = lwr, ymax = upr),
              fill = "grey70", alpha = 0.4, inherit.aes = FALSE) +
  # Fitted curve
  geom_line(data = newdata, 
            aes(x = median_genome_length, y = fit),
            color = "black", linewidth = 1.2, inherit.aes = FALSE) +
  # Data points
  geom_point(aes(color = treatment, shape = factor(age)), 
             size = 3, alpha = 0.8) +
  # Spearman correlation annotation
  annotate("text",
           x = x_pos, y = y_pos,
           label = paste0("Spearman ρ = ", round(rho, 2),
                          "\np = ", signif(pval, 2)),
           hjust = 1, vjust = 1, size = 3) +
  # Pseudo-R² annotation
  annotate("text",
           x = x_pos, y = y_pos * 0.85,
           label = paste0("Pseudo-R² = ", round(pseudo_r2, 3)),
           hjust = 1, vjust = 1, size = 3) +
  scale_color_manual(values = treatment_colours_bright, name = "Treatment") +
  scale_shape_manual(name = "DPI",
                    values = c("0" = 16, "3" = 17, "5" = 15, "7" = 18, "14" = 4)) +
  labs(x = "Median genome length (bp) per sample", y = "Richness") +
  theme_minimal() +
  custom_ggplot_theme

# Display the plot
```


