# Beta Diversity Analysis - Between Breeds

## Load Data and Plotting Utilities

```{r load_data_04E, warning=FALSE, message=FALSE}
# Load processed plotting data, color schemes, and MAG catalogue data (for genome_metadata)
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
load("data/MAG_catalogue/data.Rdata")  # Needed for genome_metadata in PCA plots
```


## Composition Barplots

This section shows the composition of bacterial orders across samples, visualized as stacked barplots.

```{r composition_barplots_04E, warning=FALSE, message=FALSE, fig.height=7, fig.width=12}
# Create stacked barplot showing order-level composition for each sample
# Colors represent different bacterial orders
tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal , y = count, fill = order)) +
  geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
  scale_fill_manual(values = order_colors[-4], drop = FALSE) +
  labs(x = "Metagenomic Samples", y = "Order composition",fill = "Read type") +
  facet_nested(. ~ animal_breed + treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        axis.text.y = element_blank(),
        legend.position = "none") +
  ggtitle("Composition of bacterial orders at 30% coverage filtered data")
```

## Layer vs Broiler Composition

This section analyzes compositional differences between layer and broiler breeds using Principal Component Analysis (PCA) and PERMANOVA statistical tests.

### PCA

#### Prepare Data for PCA

```{r pca_data_prep, warning=FALSE, message=FALSE, comment=""}
# Prepare wide format data frame for PCA analysis
# Convert from tidy format to wide format with samples as rows and genomes as columns
df_wide <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P") %>%  # Only positive samples (exclude negative controls)
  # Remove metadata columns, keep only genome counts
  select(genome, microsample, count) %>%
  # Convert to wide format: one column per genome
  pivot_wider(
    names_from = genome,   
    values_from = count,   
    values_fill = 0  # Fill missing values with 0
  ) %>%
  # Remove genomes (columns) that have zero counts in all samples
  select(where(~ any(. != 0))) %>%
  # Scale each sample to 100% (closure operation - common for compositional data)
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100 )) %>%
  # Convert microsample column to row names
  column_to_rownames(var = "microsample")

# Report zero count statistics
cat("Number of zeros:", sum(df_wide == 0), "\n")
cat("% of zeros:", sum(df_wide == 0) / length(as.matrix(df_wide)) * 100, "\n")

# Remove taxa present in fewer than 30 samples and samples with fewer than 10 taxa
# This ensures sufficient data for compositional analysis
df_wide <- remove_samples_or_taxa(
  df = df_wide,
  min_samples_per_taxon = 30,
  min_taxa_per_sample = 10
)

# Report zero count statistics after filtering
cat("Number of zeros after filtering:", sum(df_wide == 0), "\n")
cat("% of zeros after filtering:", sum(df_wide == 0) / length(as.matrix(df_wide)) * 100, "\n")

# Prepare sample metadata for PCA visualization
MACRO_sample_metadata <- sample_metadata_macro %>%
  mutate(microsample = sample) %>%
  select(microsample, animal, animal_breed, age, treatment, treatment_expl) %>%
  mutate(sample_type = "MACRO")
```

#### Perform PCA

```{r pca_perform_04E, warning=FALSE, message=FALSE, comment=""}
# Perform PCA with zero replacement, centering, scaling, and CLR transformation
# This handles compositional data appropriately
pca_result <- perform_pca(df_wide, z_delete = TRUE)
```

#### PCA Ordination Plot

```{r pca_ordination_plot, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Create PCA ordination plot showing sample composition differences
# Samples colored by breed, shaped by treatment
# Arrows (loadings) show which bacterial taxa contribute most to principal components
p <- plot_pca(
  pca_result$pca_result,
  samples_color_metadata = "animal_breed",  # Color by breed (layer vs broiler)
  samples_shape_metadata = "treatment_expl",  # Shape by treatment
  samples_color_value = c('#c1558b', "#F3B942"),  # Color scheme for breeds
  loadings_color_metadata = "order",  # Color loadings by taxonomic order
  loadings_color_value = order_colors, 
  loadings_taxon_level = "species",  # Label loadings with species names
  sample_metadata = MACRO_sample_metadata, 
  genome_metadata = genome_metadata,  # For taxonomic information on loadings
  order_colors = order_colors, 
  custom_ggplot_theme = custom_ggplot_theme,
  scaling_factor_value = 1.7,  # Scaling factor for loading arrows
  loadings_number = 17  # Number of top loadings to display
)

# Display the plot
p
```
### PERMANOVA

PERMANOVA (Permutational Multivariate Analysis of Variance) tests whether microbial community composition differs significantly between groups (breed and treatment).

#### PERMANOVA: Breed and Treatment Effects

```{r permanova_breed_treatment_prep, warning=FALSE, message=FALSE, comment=""}
# Calculate Euclidean distance matrix on CLR-transformed data
# CLR transformation makes the data suitable for Euclidean distance calculations
pca_animals_dist_matrix <- vegdist(pca_result$df_clr_dist, method = "euclidean")

# Align metadata with distance matrix samples
# Ensure filtered metadata matches samples used in PCA
pca_animals_metadata_adonis <- plot_data_stats_macro %>%
  filter(microsample %in% rownames(pca_result$df_clr_dist)) %>%
  arrange(match(microsample, rownames(pca_result$df_clr_dist))) %>%
  select(microsample, animal_breed, treatment, age)

# Verify that metadata and distance matrix samples are aligned
stopifnot(all(pca_animals_metadata_adonis$microsample == rownames(pca_result$df_clr_dist)))
```

```{r permanova_breed_treatment_test, warning=FALSE, message=FALSE, comment=""}
# Run PERMANOVA to test effects of breed and treatment on community composition
# Using 999 permutations for statistical significance testing
adonis2_result_breed_treatment <- adonis2(
  pca_animals_dist_matrix ~ animal_breed + treatment,
  data = pca_animals_metadata_adonis,
  permutations = 999,
  by = "terms"  # Test each term separately
)

# Display PERMANOVA results as formatted table
# Convert to data frame and format for display
adonis2_table <- as.data.frame(adonis2_result_breed_treatment)
knitr::kable(adonis2_table, digits = 4, caption = "PERMANOVA results: Effects of breed and treatment on beta diversity")
```

#### PERMANOVA: Breed, Treatment, and Age Effects

```{r permanova_full_model_prep, warning=FALSE, message=FALSE, comment=""}
# Recalculate distance matrix (same as above, but shown for clarity)
pca_animals_dist_matrix_full <- vegdist(pca_result$df_clr_dist, method = "euclidean")

# Use same metadata alignment
pca_animals_metadata_adonis_full <- plot_data_stats_macro %>%
  filter(microsample %in% rownames(pca_result$df_clr_dist)) %>%
  arrange(match(microsample, rownames(pca_result$df_clr_dist))) %>%
  select(microsample, animal_breed, treatment, age)

# Verify alignment
stopifnot(all(pca_animals_metadata_adonis_full$microsample == rownames(pca_result$df_clr_dist)))
```

```{r permanova_full_model_test, warning=FALSE, message=FALSE, comment=""}
# Run PERMANOVA including age as an additional factor
# This tests effects of breed, treatment, and days post-infection (age)
adonis2_result_full <- adonis2(
  pca_animals_dist_matrix_full ~ animal_breed + treatment + age,
  data = pca_animals_metadata_adonis_full,
  permutations = 999,
  by = "terms"
)

# Display PERMANOVA results as formatted table
# Convert to data frame and format for display
adonis2_table_full <- as.data.frame(adonis2_result_full)
knitr::kable(adonis2_table_full, digits = 4, caption = "PERMANOVA results: Effects of breed, treatment, and age on beta diversity")
```



### Genomes Overlap Between Breeds

This section analyzes which genomes are shared between layers and broilers, and which are unique to each breed.

```{r genomes_overlap_prep_04E, warning=FALSE, message=FALSE, comment=""}
# Identify unique genomes for each breed
# Group 1: Layers
# Group 2: Broilers
group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", animal_breed == "layer") %>%
  pull(genome) %>%
  unique()

group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", animal_breed == "broiler") %>%
  pull(genome) %>%
  unique()

venn_data <- list(
  Layers = group1,
  Broilers = group2
)

# Taxa unique to group1 only
unique_group1 <- setdiff(group1, group2) # layer
# Taxa unique to group2 only
unique_group2 <- setdiff(group2, group1) # broiler


# Filter once (layer/broiler + genomes of interest)
filtered_group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", animal_breed == "layer", genome %in% unique_group1)

filtered_group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", animal_breed == "broiler", genome %in% unique_group2)
```

```{r venn_plot_breeds_04E, message=FALSE, warning=FALSE, fig.height=4, fig.width=6}
# Create Venn diagram showing overlap of genomes between layers and broilers
ggVennDiagram(
  venn_data,
  label = "count",        
  set_color = c("Layers" = "#F3B942", "Broilers" = "#c1558b")) +
  scale_fill_gradient(low = "white", high = "grey") +
  ggtitle("Samples:") +
  coord_flip()
```

```{r venn_summary_panels_function_04E, message=FALSE, warning=FALSE, fig.height=9, fig.width=9}
# Define function to create summary panels for unique genomes per breed
plot_group_summary_panels <- function(filtered_group, order_colors,
                                      group_title = "Group", 
                                      prevalence_cutoff = 0.10,
                                      jitter_xlim = c(0, 0.3),
                                      median_xlim = c(0, 0.1)) {

  # 1. Determine denominator for prevalence
  denom <- n_distinct(filtered_group$microsample)

  # 2. Refilter unique_group_df based on prevalence cutoff
  unique_group_df <- filtered_group %>%
    group_by(species) %>%
    summarise(
      median = median(count, na.rm = TRUE),
      n_samples = n_distinct(microsample),
      prevalence = n_samples / denom,
      .groups = "drop"
    ) %>%
    filter(prevalence > prevalence_cutoff)

  # 3. Prepare plot_df for jitter plot
  plot_df <- filtered_group %>%
    inner_join(unique_group_df %>% select(species, prevalence), by = "species") %>%
    mutate(species = fct_reorder(species, prevalence, .desc = FALSE))

  # 4. Jitter plot (p1)
  p1 <- ggplot(plot_df, aes(x = count, y = species, color = order)) +
    geom_jitter(alpha = 0.7, size = 0.9) +
    scale_color_manual(values = order_colors) +
    theme_minimal() +
    custom_ggplot_theme +
    theme(
      plot.margin = margin(0.1, 0.1, 0.1, 0.1),
      axis.text.x = element_text(size = 6)
    ) +
    scale_x_continuous(limits = jitter_xlim) +
    labs(y = "Species", x = "Relative abundance") +
    ggtitle(group_title)

  # 5. Prevalence plot (p2)
  p2 <- unique_group_df %>%
    mutate(species = fct_reorder(species, prevalence, .desc = FALSE)) %>%
    ggplot(aes(x = prevalence, y = species)) +
    geom_col() +
    geom_text(aes(label = round(prevalence, 2), x = prevalence + 0.02),
              size = 3, hjust = 0) +
    labs(y = "Species", x = "Prevalence") +
    theme_minimal() +
    custom_ggplot_theme +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      plot.margin = margin(0.1, 0.1, 0.1, 0.1),
      axis.text.x = element_text(size = 6)
    ) +
    scale_x_continuous(limits = c(0, 1))

  # 6. Median plot (p3)
  p3 <- unique_group_df %>%
    mutate(species = fct_reorder(species, prevalence, .desc = FALSE)) %>%
    ggplot(aes(x = median, y = species)) +
    geom_col() +
    geom_text(aes(label = round(median, 3), x = 0.075),
              size = 3, hjust = 0) +
    labs(y = "Species", x = "Median Rel. abun.") +
    theme_minimal() +
    custom_ggplot_theme +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.minor.x = element_blank(),
      plot.margin = margin(0.1, 0.1, 0.1, 0.1),
      axis.text.x = element_text(size = 6)
    ) +
    scale_x_continuous(limits = median_xlim)

  # 7. Combine plots
  final_plot <- (p1 + p3 + p2) +
    plot_layout(widths = c(2, 1, 1), guides = "collect") &
    theme(legend.position = "bottom", legend.box = "vertical")

  return(final_plot)
}
```

```{r venn_unique_layer_genomes_04E, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
plot_unique_broilers <- plot_group_summary_panels(
  filtered_group = filtered_group2,
  order_colors = order_colors,
  group_title = "Broilers"
)

plot_unique_layers <- plot_group_summary_panels(
  filtered_group = filtered_group1,
  order_colors = order_colors,
  group_title = "Layers"
)
```

```{r venn_unique_broiler_genomes_04E, message=FALSE, warning=FALSE, fig.height=10, fig.width=9}
plot_unique_broilers / plot_unique_layers  + plot_layout(heights = c(4, 1), guides = "collect") &
  theme(legend.position = "bottom", legend.box = "vertical")
```

## ALDEx2 Differential Abundance Analysis

ALDEx2 (Analysis of Differential Abundance taking sample variation into account) is used to identify bacterial taxa that differ significantly in abundance between breeds.

### Prepare Data for ALDEx2

```{r aldex2_data_prep_04E, warning=FALSE, message=FALSE, comment=""}
# Prepare wide format data frame for ALDEx2 analysis
# ALDEx2 requires samples as columns and taxa as rows
df_wide <- tidy_plot_c_read_counts_macro_filt_30 %>%
  filter(type_simple == "P") %>%  # Only positive samples
  # Remove metadata, keep only genome counts
  select(genome, microsample, count) %>%
  # Convert to wide format: one column per sample
  pivot_wider(
    names_from = genome,   
    values_from = count,   
    values_fill = 0  # Fill missing values with 0
  ) %>%
  # Remove genomes with zero counts in all samples
  select(where(~ any(. != 0))) %>%
  # Convert microsample column to row names (samples as rows)
  column_to_rownames(var = "microsample")

# Report zero count statistics
cat("Number of zeros:", sum(df_wide == 0), "\n")
cat("% of zeros:", sum(df_wide == 0) / length(as.matrix(df_wide)) * 100, "\n")

# Filter: remove taxa present in fewer than 30 samples and samples with fewer than 10 taxa
df_wide <- remove_samples_or_taxa(
  df = df_wide,
  min_samples_per_taxon = 30,
  min_taxa_per_sample = 10
)

# Report zero count statistics after filtering
cat("Number of zeros after filtering:", sum(df_wide == 0), "\n")
cat("% of zeros after filtering:", sum(df_wide == 0) / length(as.matrix(df_wide)) * 100, "\n")

# Count samples per breed group for ALDEx2
df_wide_groups <- sample_metadata_macro %>%
  filter(sample %in% rownames(df_wide)) %>%
  count(animal_breed)

# Display group sizes as formatted table
knitr::kable(df_wide_groups, caption = "Number of samples per breed group for ALDEx2 analysis")

# Create condition vector for ALDEx2 (breed labels for each sample)
groups_conds <- unlist(mapply(rep, df_wide_groups$animal_breed, df_wide_groups$n), use.names = FALSE)

# Transpose data frame for ALDEx2 (taxa as rows, samples as columns)
df_wide_scaled_T <- as.data.frame(t(df_wide))
```

### Run ALDEx2 Analysis

```{r aldex2_run_04E, warning=FALSE, message=FALSE, comment=""}
# Run ALDEx2 with Welch's t-test
# mc.samples: number of Monte Carlo samples for posterior distribution
set.seed(1)
treatments_ALDEx2.all <- aldex(
  df_wide_scaled_T, 
  groups_conds, 
  mc.samples = 128, 
  test = "t",  # Welch's t-test
  effect = TRUE,
  include.sample.summary = TRUE, 
  denom = "all",  # Use all taxa as denominator
  verbose = FALSE
)

# Extract relevant ALDEx2 results for plotting and further analysis
treatments_ALDEx2_subset <- treatments_ALDEx2.all %>%
  select(diff.btw, diff.win, effect, overlap,
         we.ep, we.eBH, wi.ep, wi.eBH) %>%
  mutate(genome = rownames(.)) %>%  # Add genome names as column
  relocate(genome, .before = 1)
```

### ALDEx2 Diagnostic Plots

```{r aldex2_diagnostic_plots_04E, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Create diagnostic plots for ALDEx2 results
par(mfrow = c(2, 2))  # 2x2 plot layout

# MA plot (Mean vs. Difference)
aldex.plot(treatments_ALDEx2.all, type = "MA", test = "welch")

# MW plot (Median vs. Within-group difference)
aldex.plot(treatments_ALDEx2.all, type = "MW", test = "welch")

# Effect size vs. adjusted p-value plot
plot(treatments_ALDEx2.all$effect, treatments_ALDEx2.all$we.eBH, 
     log = "y", pch = 19, main = "Effect", cex = 0.5, 
     xlab = "Effect size", ylab = "Expected Benjamini-Hochberg P")
abline(h = 0.05, lty = 2)  # Significance threshold

# Volcano plot (Difference vs. adjusted p-value)
plot(treatments_ALDEx2.all$diff.btw, treatments_ALDEx2.all$we.eBH, 
     log = "y", pch = 19, main = "Volcano", cex = 0.5, 
     xlab = "Difference", ylab = "Expected Benjamini-Hochberg P")
abline(h = 0.05, lty = 2)  # Significance threshold
```


### ALDEx2 Significant Taxa Visualization

```{r aldex2_significant_taxa_plot_04E, warning=FALSE, message=FALSE, fig.height=3.2, fig.width=9}
# Prepare ALDEx2 results for visualization
# Filter for significantly different taxa (we.eBH <= 0.01) with meaningful effect sizes
ALDEx2_significant <- treatments_ALDEx2_subset %>%
  # Filter ALDEx2 taxa for significance by the welch test (wilcoxon found more hits)
  filter(we.eBH <= 0.05) %>% # 0.005
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  # Filter so that the difference between groups is more than x
  filter(abs(diff.btw) >= 0.1) %>%
  arrange(diff.btw) %>%
  mutate(
    genome_label = paste0(genome, " (", species, ", ", phylum, ")"),
    genome_label = factor(genome_label, levels = genome_label[order(diff.btw)]),
    color_group = ifelse(abs(effect) > 1, "Larger than 1", "Smaller than 1")
  )

# Prepare tidy df with rel. abun. for plotting
ALDEx2_rel_abun_tidy <- as.data.frame(df_wide / rowSums(df_wide)) %>% # Close to 1
  rownames_to_column("microsample") %>%
  pivot_longer(
    cols = -microsample,
    names_to = "genome",
    values_to = "relabun"
  ) %>%
  filter(relabun > 0) %>%
  left_join(genome_metadata, by = "genome") %>%
  left_join(plot_data_stats_macro, by = "microsample") %>%
  left_join(treatments_ALDEx2_subset, by = "genome") %>%
  mutate(
    phylum = factor(phylum, levels = unique(genome_metadata$phylum)),
    order = factor(order, levels = unique(genome_metadata$order))
  )


# ALDEx2 differential abundance barplot (using fill = effect size group)
treatment_diff_abun_barplot <- ggplot(ALDEx2_significant, aes(x = diff.btw, y = genome_label)) +
  geom_col(aes(fill = color_group), position = "identity", width = 0.7) +
  scale_fill_manual(values = c("Larger than 1" = "#e56969", "Smaller than 1" = "gray")) +
  labs(
    y = NULL, # "Species"
    x = "diff.btw",
    fill = "Effect Size"
  ) +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.y = element_blank()) +
  theme(axis.text.y = element_blank()) +
  theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5))


# Order species by median diff.btw for plot
treatment_summary <- ALDEx2_rel_abun_tidy %>%
  group_by(species) %>%
  summarise(diff = median(diff.btw, na.rm = TRUE)) %>%
  arrange(-diff)

# Plot rel. abund. for species in ALDEx2 barplot
 
treatment_relat_abun_jitterplot_left <- ALDEx2_rel_abun_tidy %>%
  filter(animal_breed == "broiler") %>%
  filter(genome %in% ALDEx2_significant$genome) %>%
  mutate(species = factor(species, levels = rev(treatment_summary %>% pull(species)))) %>%
  ggplot(aes(x = relabun, y = species, group = species, color = order)) +
  scale_color_manual(values = order_colors) +
  geom_jitter(alpha = 0.7, size= 0.9) +
  facet_nested(. ~ animal_breed, scales = "free", space = "free", switch = "y") +
  theme_minimal() +
  custom_ggplot_theme +
  scale_x_continuous(limits = c(0, 1)) +
  theme(panel.spacing = unit(1, "lines")) +
  theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5))

treatment_relat_abun_jitterplot_right <- ALDEx2_rel_abun_tidy %>%
  filter(animal_breed == "layer") %>%
  filter(genome %in% ALDEx2_significant$genome) %>%
  mutate(species = factor(species, levels = rev(treatment_summary %>% pull(species)))) %>%
  ggplot(aes(x = relabun, y = species, group = species, color = order)) +
  labs(x="Relative abun", y = NULL) +
  scale_color_manual(values = order_colors) +
  geom_jitter(alpha = 0.7, size= 0.9) +
  facet_nested(. ~ animal_breed, scales = "free", space = "free", switch = "y") +
  theme_minimal() +
  custom_ggplot_theme +
  scale_x_continuous(limits = c(0, 1)) +
  theme(panel.spacing = unit(1, "lines")) +
  theme(axis.text.y = element_blank()) +
  theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5))


treatment_all_plots <- treatment_relat_abun_jitterplot_left + treatment_diff_abun_barplot + treatment_relat_abun_jitterplot_right +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom", legend.box = "vertical")

treatment_all_plots
```





## ANCOM Differential Abundance Analysis

ANCOM (Analysis of Composition of Microbiomes) is an alternative method to identify differentially abundant taxa between groups. ANCOMBC2 is used here.

### Prepare Data for ANCOM

```{r ancom_data_prep_04E, warning=FALSE, message=FALSE, comment=""}
df_wide <- tidy_plot_c_read_counts_macro_filt_30 %>%
  filter(type_simple == "P") %>%
  # Remove metadata
  select(genome, microsample, count) %>%
  # Make tidy df into wide df
  pivot_wider(
    names_from = genome,   
    values_from = count,   
    # Fill missing values with 0
    values_fill = 0) %>%
  # Remove taxa (columns) that are zero for all leftover samples (rows)
  select(where(~ any(. != 0))) %>%
  # turn the column 'microsample'into the name of the rows of the df
  column_to_rownames(var = "microsample")

cat("Number of zeros:", sum(df_wide == 0), "\n")
cat("% of zeros:", sum(df_wide == 0) / length(as.matrix(df_wide)) * 100, "\n")

# Remove taxa with less than 0.1% rel. abun. in all samples
# df_wide <- df_wide %>% select(where(~ any(. >= 0.1)))

# Remove taxa in few samples, and samples with few taxa
df_wide <- remove_samples_or_taxa(df = df_wide,
                                     min_samples_per_taxon = 30,
                                     min_taxa_per_sample = 10)
cat("Number of zeros:", sum(df_wide == 0), "\n")
cat("% of zeros:", sum(df_wide == 0) / length(as.matrix(df_wide)) * 100, "\n")


# 1. Response matrix (counts): samples as rows, genomes as columns
counts_df <- t(df_wide)  # already samples x genomes

metadata_df <- sample_metadata_macro %>%
  filter(sample %in% colnames(counts_df)) %>%
  column_to_rownames("sample")

counts_df <- counts_df[, rownames(metadata_df)]

# Create phyloseq object
otu_table <- otu_table(as.matrix(counts_df), taxa_are_rows = TRUE)
sample_data <- sample_data(metadata_df)

ps_obj <- phyloseq(otu_table, sample_data)

# Step 2: Run ANCOMBC2 ----

out <- ANCOMBC::ancombc2(
  data        = ps_obj,
  fix_formula = "animal_breed + treatment + age",
  rand_formula = NULL,     # or e.g. "~ 1 | animal_id" if you have repeated measures
  p_adj_method = "BH",
  prv_cut      = 0.1,
  lib_cut      = 1000,
  group        = "animal_breed",
  struc_zero   = TRUE, #TRUE
  neg_lb       = TRUE
)

# Step 3: Extract results ----

res_df <- out$res   # Differential abundance results
```


```{r ancom_lfc_interpretation_04E, warning=FALSE, message=FALSE, comment=""}
# Common |log2 fold change| thresholds for biological interpretation:
# |lfc| ≥ 0.58   → ~1.5× change  (small but possibly meaningful)
# |lfc| ≥ 1.0    → 2× change     (moderate effect, often used cutoff)
# |lfc| ≥ 1.5    → ~2.8× change  (strong effect)
# |lfc| ≥ 2.0    → 4× change     (very strong effect, highly robust)


sig_df <- res_df %>%
  mutate(significant = q_animal_breedlayer < 0.05) %>%
  filter(abs(lfc_animal_breedlayer) >= 0.58) %>%
  rename(genome=taxon)%>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  arrange(lfc_animal_breedlayer) 

sig_df <- sig_df %>%
  mutate(
    genome_label = paste0(genome, " (", species, ", ", phylum, ")"),
    genome_label = factor(genome_label, levels = genome_label[order(lfc_animal_breedlayer)]),
    species_label = species,
    species_label = factor(species_label, levels = species_label[order(lfc_animal_breedlayer)]))

# ALDEx2 differential abundance barplot (using fill = effect size group)
treatment_diff_abun_barplot <- ggplot(sig_df, 
       aes(x = lfc_animal_breedlayer, y = species_label, fill = order)) +
  geom_col(position = "identity", width = 0.7) +
  scale_fill_manual(values = order_colors) +
  labs(y = "Species", x = "Log2-fold change" ) +
  theme_minimal() +
  custom_ggplot_theme +
  theme(plot.margin = margin(0.1, 0.1, 0.1, 0.1))

# treatment_diff_abun_barplot
```


```{r ancom_relative_abundance_prep_04E, warning=FALSE, message=FALSE, comment=""}
# Prepare tidy data frame with relative abundance for plotting
diff_abun_rel_abun_tidy <- as.data.frame(df_wide / rowSums(df_wide)) %>% # Close to 1
  rownames_to_column("microsample") %>%
  pivot_longer(
    cols = -microsample,
    names_to = "genome",
    values_to = "relabun"
  ) %>%
  filter(relabun > 0) %>%
  # left_join(genome_metadata, by = "genome") %>%
  left_join(plot_data_stats_macro, by = "microsample") %>%
  left_join(sig_df, by = "genome") %>%
  mutate(
    phylum = factor(phylum, levels = unique(genome_metadata$phylum)),
    order = factor(order, levels = unique(genome_metadata$order))
  )

# Order species by median diff.btw for plot
diff_abun_rel_abun_tidy_summary <- diff_abun_rel_abun_tidy %>%
  group_by(species) %>%
  summarise(diff = median(lfc_animal_breedlayer, na.rm = TRUE)) %>%
  arrange(-diff)

# Plot rel. abund. for species in ALDEx2 barplot
 
treatment_relat_abun_jitterplot <- diff_abun_rel_abun_tidy %>%
  filter(genome %in% sig_df$genome) %>%
  mutate(species = factor(species, levels = rev(diff_abun_rel_abun_tidy_summary %>% pull(species)))) %>%
  ggplot(aes(x = relabun, y = species, group = species, color = order)) +
  scale_color_manual(values = order_colors) +
  geom_jitter(alpha = 0.7, size= 0.9) +
  facet_nested(. ~ animal_breed, scales = "free", space = "free", switch = "y") +
  theme_minimal() +
  custom_ggplot_theme +
  scale_x_continuous(limits = c(0, 0.6)) +
  theme(panel.spacing = unit(1, "lines")) +
  theme(plot.margin = unit(c(0.5, 0, 0, 0), "cm"))
# treatment_relat_abun_jitterplot
```

```{r ancom_significant_taxa_plot_04E, warning=FALSE, message=FALSE, fig.height=5, fig.width=10}
# Create plot showing significant taxa from ANCOM analysis
plot_1 <- ggplot(sig_df, 
       aes(x = lfc_animal_breedlayer, y = species_label, fill = order)) +
  geom_col(position = "identity", width = 0.7) +
  scale_fill_manual(values = order_colors) +
  labs(y = NULL, x = "Log2-fold change" ) +
  theme_minimal() +
  custom_ggplot_theme +
  theme(axis.text.y = element_blank()) +
  theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5))

plot_left <- diff_abun_rel_abun_tidy %>%
  filter(animal_breed == "broiler") %>%
  filter(genome %in% sig_df$genome) %>%
  mutate(species = factor(species, levels = rev(diff_abun_rel_abun_tidy_summary %>% pull(species)))) %>%
  ggplot(aes(x = relabun, y = species, group = species, color = order)) +
  scale_color_manual(values = order_colors) +
  geom_jitter(alpha = 0.7, size= 0.9) +
  facet_nested(. ~ animal_breed, scales = "free", space = "free", switch = "y") +
  theme_minimal() +
  custom_ggplot_theme +
  scale_x_continuous(limits = c(0, 0.6)) +
  theme(panel.spacing = unit(1, "lines")) +
  theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5))

plot_right <- diff_abun_rel_abun_tidy %>%
  filter(animal_breed == "layer") %>%
  filter(genome %in% sig_df$genome) %>%
  mutate(species = factor(species, levels = rev(diff_abun_rel_abun_tidy_summary %>% pull(species)))) %>%
  ggplot(aes(x = relabun, y = species, group = species, color = order)) +
  scale_color_manual(values = order_colors) +
  geom_jitter(alpha = 0.7, size= 0.9) +
  facet_nested(. ~ animal_breed, scales = "free", space = "free", switch = "y") +
  labs(x="Relative abun", y = NULL) +
  theme_minimal() +
  custom_ggplot_theme +
  scale_x_continuous(limits = c(0, 0.6)) +
  theme(panel.spacing = unit(1, "lines")) +
  theme(axis.text.y = element_blank()) +
  theme(plot.margin = margin(0.5, 0.5, 0.5, 0.5))

plot_left + plot_1 + plot_right + plot_layout(guides = "collect") &
  theme(legend.position = "bottom", legend.box = "vertical")
```
  


### Venn Diagram: Control Treatment (TF3) Only

```{r venn_tf3_prep_04E, warning=FALSE, message=FALSE, comment=""}
group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF3") %>%
  filter(type_simple == "P", animal_breed == "layer") %>%
  pull(genome) %>%
  unique()

group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF3") %>%
  filter(type_simple == "P", animal_breed == "broiler") %>%
  pull(genome) %>%
  unique()

venn_data <- list(
  Layers = group1,
  Broilers = group2
)

# Taxa unique to group1 only
unique_group1 <- setdiff(group1, group2) # layer
# Taxa unique to group2 only
unique_group2 <- setdiff(group2, group1) # broiler


# Filter once (layer/broiler + genomes of interest)
filtered_group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", animal_breed == "layer", genome %in% unique_group1)

filtered_group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(type_simple == "P", animal_breed == "broiler", genome %in% unique_group2)
```

```{r venn_tf3_all_groups_04E, message=FALSE, warning=FALSE, fig.height=3.5, fig.width=5}

ggVennDiagram(
  venn_data,
  label = "count",        
  set_color = c("Layers" = "#F3B942", "Broilers" = "#c1558b")) +
  scale_fill_gradient(low = "white", high = "grey") +
  ggtitle("Samples from TF1 only:") +
  coord_flip()
```

### Venn of all groups separately
```{r venn_all_groups_prep_04E, warning=FALSE, message=FALSE, comment=""}
# Prepare data for Venn diagrams comparing all treatment groups separately
group1_TF3 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF3") %>%
  filter(type_simple == "P", animal_breed == "layer") %>%
  pull(genome) %>%
  unique()

group1_TF2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF2") %>%
  filter(type_simple == "P", animal_breed == "layer") %>%
  pull(genome) %>%
  unique()

group1_TF1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF1") %>%
  filter(type_simple == "P", animal_breed == "layer") %>%
  pull(genome) %>%
  unique()

group2_TF3 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF3") %>%
  filter(type_simple == "P", animal_breed == "broiler") %>%
  pull(genome) %>%
  unique()


group2_TF2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF2") %>%
  filter(type_simple == "P", animal_breed == "broiler") %>%
  pull(genome) %>%
  unique()

group2_TF1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment == "TF1") %>%
  filter(type_simple == "P", animal_breed == "broiler") %>%
  pull(genome) %>%
  unique()

venn_data <- list(
  Layers_TF3 = group1_TF3,
  Layers_TF2 = group1_TF2,
  Layers_TF1 = group1_TF1,
  Broilers_TF3 = group2_TF3,
  Broilers_TF2 = group2_TF2,
  Broilers_TF1 = group2_TF1
)

# # Taxa unique to group1 only
# unique_group1 <- setdiff(group1, group2) # layer
# # Taxa unique to group2 only
# unique_group2 <- setdiff(group2, group1) # broiler
# 
# 
# # Filter once (layer/broiler + genomes of interest)
# filtered_group1 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
#   filter(type_simple == "P", animal_breed == "layer", genome %in% unique_group1)
# 
# filtered_group2 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
#   filter(type_simple == "P", animal_breed == "broiler", genome %in% unique_group2)
```

```{r venn_all_treatments_combined_04E, message=FALSE, warning=FALSE, fig.height=10, fig.width=15}

ggVennDiagram(
  venn_data,
  label = "count") +
  scale_fill_gradient(low = "white", high = "grey") +
  ggtitle("Samples:") +
  coord_flip()
```










