# Diversity Partitioning Analysis

## Overview

This chapter performs hierarchical additive partitioning of alpha diversity to quantify how diversity is distributed across different hierarchical levels (within and among animals, treatments, and breeds).

**Note:** This analysis requires specific functions for hierarchical partitioning that are defined within this chapter.

## Load Data and Plotting Utilities

```{r load_data_04G, warning=FALSE, message=FALSE}
# Load processed plotting data and color schemes
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
```

## Hierarchical Diversity Partitioning Functions

These functions are used to calculate and visualize hierarchical additive partitioning of diversity.

```{r diversity_partitioning_functions_04G, warning=FALSE, message=FALSE, comment=""}
# Function to calculate hierarchical additive partitioning

# Function to calculate hierarchical additive partitioning
hier_adipart <- function(comm, hier, nsim = 999) {
  comm_bin <- (as.matrix(comm) > 0) * 1
  alpha_obs <- mean(rowSums(comm_bin))

  get_level_richness <- function(fac) {
    tapply(seq_len(nrow(comm_bin)), fac, function(idx) {
      sum(colSums(comm_bin[idx, , drop = FALSE]) > 0)
    }) |> mean()
  }

  level_names <- names(hier)
  level_rich  <- sapply(hier, get_level_richness)
  gamma_obs   <- sum(colSums(comm_bin) > 0)

  # exactly L betas
  betas <- c(level_rich[1] - alpha_obs, diff(level_rich))
  names(betas) <- paste0("beta (", level_names, ")")
  observed <- c(alpha = alpha_obs, betas, gamma = gamma_obs)

  stat_fun <- function(x) {
    x <- (x > 0) * 1
    alpha <- mean(rowSums(x))
    lvl <- sapply(hier, function(fac) {
      tapply(seq_len(nrow(x)), fac, function(idx)
        sum(colSums(x[idx, , drop = FALSE]) > 0)) |> mean()
    })
    gamma <- sum(colSums(x) > 0)
    betas <- c(lvl[1] - alpha, diff(lvl))     # ✅ no gamma - last_level
    names(betas) <- paste0("beta (", level_names, ")")
    c(alpha = alpha, betas, gamma = gamma)
  }

  sim <- oecosimu(comm_bin, stat_fun, method = "r2dtable", nsimul = nsim)

  tibble::tibble(
    component = names(observed),
    observed  = as.numeric(observed),
    expected  = as.numeric(sim$oecosimu$mean),
    pval      = as.numeric(sim$oecosimu$pval)
  )
}




plot_hier_adipart <- function(res, title = NULL) {
  p_to_stars <- function(p) case_when(
    p <= 0.001 ~ "***",
    p <= 0.01  ~ "**",
    p <= 0.05  ~ "*",
    TRUE      ~ ""
  )
  
  res_long <- res %>%
    pivot_longer(c(observed, expected), names_to = "type", values_to = "value") %>%
    mutate(
    component = factor(
      component,
      levels = c("alpha",
                 "beta (animal)",
                 "beta (treatment)",
                 "beta (animal_breed)",
                 "gamma")
    )
  )
  
  ggplot(res_long, aes(x = component, y = value, fill = type)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "black") +
    geom_text(data = res, aes(x = component,
                              y = pmax(observed, expected) * 1.05,
                              label = p_to_stars(pval)),
              inherit.aes = FALSE, vjust = 0) +
    labs(x = NULL, y = "Richness", fill = NULL, title = title) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 15, hjust = 1),
          legend.position = "top")
}

```


## Prepare Data for Diversity Partitioning

This section prepares the community matrix and hierarchical metadata structure required for diversity partitioning analysis.

```{r diversity_partitioning_data_prep_04G, warning = FALSE, message = FALSE, comment=""}
# Filter metadata for positive samples retained after filtering
metadata <- plot_data_stats_macro %>%
  filter(
    type_simple %in% c("P"),
    filter_status %in% c("Retained by filtering"))
    
comm_data <- genome_counts_macro_filt_30_zerosrem %>%
  select(genome, metadata$microsample) %>%
  t() %>%
  data.frame() %>%
  row_to_names(row_number = 1) %>%
  mutate(across(everything(), as.numeric))

# table(rowSums(comm_data) > 0)
# all(metadata$microsample == rownames(comm_data))
# table(is.na(comm_data))
# table(metadata$section, metadata$cryosection)

comm_ca <- comm_data[, colSums(comm_data) > 0]

metadata_ca <- metadata
# table(metadata$animal, metadata$age)
# table(metadata$treatment, metadata$animal_breed)

# Ensure hierarchy variables are factors, in order: cryosection (lowest), animal, treatment (highest)
hier_ca <- metadata_ca %>%
  transmute(
    animal = factor(animal),
    treatment   = factor(treatment),
    animal_breed   = factor(animal_breed)
  )
```


## Diversity Partitioning Analysis

This section performs the hierarchical additive partitioning analysis and visualizes the results.

```{r diversity_partitioning_analysis_04G, warning=FALSE, message=FALSE, comment=""}
# Perform hierarchical additive partitioning
# comm_ca is the samples x genomes matrix
# and hier_ca is your data frame with animal, treatment, animal_breed (in order lowest → highest)
res_ca <- hier_adipart(comm_ca, hier_ca, nsim = 999)

# Display diversity partitioning results as formatted table
# The hier_adipart function returns a tibble with columns: component, observed, expected, pval
# Component structure: alpha, beta(animal), beta(treatment), beta(animal_breed), gamma
# Map component names to more descriptive labels for display
component_labels <- c(
  "alpha" = "Within animals (alpha diversity)",
  "beta (animal)" = "Among animals, within treatments",
  "beta (treatment)" = "Among treatments, within breeds",
  "beta (animal_breed)" = "Among breeds",
  "gamma" = "Total diversity (gamma)"
)

# Create formatted data frame for display
# Define the desired order first
component_order <- c("alpha", "beta (animal)", "beta (treatment)", "beta (animal_breed)", "gamma")

res_ca_df <- res_ca %>%
  mutate(
    Component = component_labels[component],  # Map to descriptive labels
    Diversity = observed,
    Expected = expected,
    P_value = pval,
    order = match(component, component_order)  # Add ordering column
  ) %>%
  arrange(order) %>%  # Arrange before selecting
  select(Component, Diversity, Expected, P_value)  # Select final columns

# Display as formatted table
knitr::kable(res_ca_df, digits = 4, caption = "Diversity partitioning results (additive partitioning of alpha diversity)")
```

```{r diversity_partitioning_plot_04G, warning=FALSE, message=FALSE, fig.height=6, fig.width=7}
# Create visualization of diversity partitioning results
plot_hier_adipart(res_ca, title = "Caecum") +
  custom_ggplot_theme
```
