# PacBio Selection Analysis

## Overview

This chapter identifies samples suitable for PacBio sequencing based on DNA concentration, elution volume, and alpha diversity metrics. PacBio long-read sequencing requires high-quality DNA samples with sufficient quantity and concentration for library preparation.

## Selection Criteria

PacBio sequencing requirements:
- **Elution volume**: >50 uL
- **DNA concentration**: >80 ng/uL
- **Minimum total DNA**: 50 × 80 = 4000 ng per sample

**Selection strategy:**
- Target: 24 samples to provide 4800 ng total DNA (200 ng per sample × 24)
- Taxonomic representation: 2 breeds × 3 treatments = 6 treatment groups
- Temporal coverage: 4 time points (days 0, 5, 7, 14 post-infection)
- Total samples: 4 days × 6 groups = 24 samples

## Load Data

```{r load_data_05A, eval=TRUE, message=FALSE, warning=FALSE}
# Load processed plotting data, color schemes, and extract concentration data
# Note: extract_concentration_data.Rdata is created in this chapter if not already present
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
load("data/macro/extract_concentration_data.Rdata")
```

## Prepare Extract Metadata

This section loads DNA extraction metadata from Airtable, including concentration, elution volume, and remaining quantities. This metadata is used to identify samples meeting PacBio requirements.

```{r load_extract_metadata_05A, eval=FALSE, message=FALSE}
# Load DNA extraction metadata from Airtable
# This includes concentration, elution volume, and leftover quantities for each extract
extract_metadata_macro <- airtable("2-EX-Info", "apphvsV2YKhTCMKGS") %>% 
  # select columns to extract
  read_airtable(., fields = c(
    "Extract", "Barcode", "Concentration_ng_uL", "ElutionVolume_uL",
    "LeftOver_uL", "LeftOver_ng", "AcidNucleicMaterial",
    "LabBatch_text", "ExperimentalUnit"), id_to_col = TRUE) %>%
  # rename columns
  rename(
    EX_batch = "LabBatch_text",
    extract = "Extract",
    animal = "ExperimentalUnit",
    tube_barcode = "Barcode",
    DNA_concentration = "Concentration_ng_uL",
    elution_uL = "ElutionVolume_uL",
    leftover_uL = "LeftOver_uL",
    leftover_ng = "LeftOver_ng",
    nucleic_acid = "AcidNucleicMaterial"
  ) %>%
    # select relevant batches
  filter(EX_batch %in% c("EXB020","EXB021")) %>%
  unnest(c(nucleic_acid, animal)) %>%
  # select relevant columns
  select(EX_batch, extract, animal, tube_barcode, DNA_concentration, elution_uL, leftover_uL, leftover_ng, nucleic_acid) %>%
  filter(nucleic_acid == "DNA") %>%
  # un-nest columns that appeared as a list
  arrange(animal)
```      


## Merge Metadata with Sample Information

This section merges extract metadata with sample metadata and alpha diversity metrics to create a comprehensive dataset for sample selection.

```{r merge_extract_metadata_05A, eval=FALSE, message=FALSE}
# Merge extract metadata with sample metadata
# This links extraction information to experimental samples (animal, treatment, breed, age)
extract_sample_metadata_macro <- extract_metadata_macro %>% 
  left_join(sample_metadata_macro, by = join_by(animal == animal)) %>%
  rename(microsample=sample)

extract_sample_div_metadata_macro <- extract_sample_metadata_macro %>%
 left_join(alpha_div_macro_filtered_30, by = join_by(microsample == microsample)) 
```

## Visualize DNA Concentration and Richness

This section creates visualizations to assess DNA extraction quality (leftover quantity) and taxonomic richness across all samples, helping identify candidates for PacBio sequencing.

```{r pacbio_barplots_05A, eval=TRUE, message=FALSE, fig.height=6, fig.width=15}
# Plot 1: DNA leftover quantity by extract
# Shows remaining DNA quantity after initial use, indicating sample availability
barplot_1 <- extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = extract, y = leftover_ng)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "DNA extracts", y = "leftover_ng", fill = "Read type") +
  facet_nested(. ~ animal_breed + treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(), #axis.text.x = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers + Layers")
# barplot_1  # Individual plot commented out - used in combined plot


# Plot 2: Taxonomic richness by animal, colored by leftover DNA quantity
# Combines diversity metrics with DNA availability information
barplot_2 <- extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal, y = richness, fill = leftover_ng)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "DNA extracts", y = "richness", fill = "Read type") +
  scale_fill_gradient(low = "#c90076", high = "#3598bf", name = "leftover ng") +
  facet_nested(. ~ animal_breed + treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(), #axis.text.x = element_blank(),
        legend.position = "top",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers + Layers")
# barplot_2  # Individual plot commented out - used in combined plot


# Plot 3: Taxonomic composition by order
# Shows community structure to ensure taxonomic diversity in selected samples
barplot_3 <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal , y = count, fill = order)) +
  geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
  scale_fill_manual(values = order_colors[-4], drop = FALSE) +
  labs(x = "Composition",y = "Microsamples",fill = "Read type") +
  facet_nested(. ~ animal_breed + treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_blank(),
        legend.position = "none") +
  ggtitle("Composition of bacterial orders - 30% filt")
# barplot_3  # Individual plot commented out - used in combined plot
```







```{r pacbio_barplots_combined_05A, eval=TRUE, message=FALSE, fig.height=15, fig.width=18}
# Combined visualization: richness and taxonomic composition
# Shows both diversity metrics and community structure for sample assessment
barplot_2/barplot_3
```

## Identify Best Samples

This section applies selection criteria to identify the best samples for PacBio sequencing, ranking samples by combined richness and DNA availability metrics.

```{r pacbio_selection_05A, eval=TRUE, message=FALSE, fig.height=15, fig.width=18}
# Define required DNA quantity per sample for PacBio library preparation
required_ng_PacBio = 220 

# Select best samples based on combined ranking of richness and DNA availability
# Filter for target time points (days 0, 5, 7, 14) and exclude TF0 control
best_samples <- extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  filter(age %in% c(0, 5, 7, 14)) %>%
  group_by(animal_breed, treatment_expl, age) %>%
  mutate(
    richness_rank = rank(desc(richness), ties.method = "min"),
    leftover_rank = rank(desc(leftover_ng), ties.method = "min"),
    combined_rank = richness_rank*1.2 + leftover_rank
  ) %>%
  slice_min(combined_rank, n = 1, with_ties = FALSE) %>%
  ungroup()

# Calculate volume and quantity requirements for PacBio sequencing
# Determines how much volume is needed per sample and what remains after extraction
best_samples <- best_samples %>%
  mutate(
    uL_required = ceiling(required_ng_PacBio / DNA_concentration),  # Volume needed for 220 ng
    uL_left_after_pacbio = (leftover_uL - uL_required),            # Remaining volume
    ng_left_after_pacbio = (leftover_ng - required_ng_PacBio)      # Remaining quantity
  )
```  

## Visualize Selected Samples

This section marks selected samples on the visualization plots to show which samples were chosen for PacBio sequencing.

```{r pacbio_barplots_marked_05A, eval=TRUE, message=FALSE, fig.height=13, fig.width=20}
# Mark selected samples with star symbols on the richness plot
barplot_4 <- barplot_2 +
  geom_point(
    data = best_samples,
    aes(x = animal, y = richness),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

# Mark selected samples on the composition plot
barplot_5 <- barplot_3 +
  geom_point(
    data = best_samples,
    aes(x = animal, y = 1),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

# Display combined plot with selected samples marked
barplot_4/barplot_5

```


## Calculate Pooling Requirements

This section calculates the pooling requirements for combining selected samples, including total volume, concentration, and condensation needs.

```{r pacbio_pool_calc_05A, eval=TRUE, message=FALSE, fig.height=13, fig.width=20}
# Display selected samples summary
best_samples

# Calculate pooling statistics


total_uL <- sum(best_samples$uL_required, na.rm = TRUE)
total_ng <- required_ng_PacBio * 24
pool_conc <- round(total_ng / total_uL, 2)
target_conc <- 80
condensed_volume <- round(total_ng / target_conc, 2)



cat(
  "The pool of 24 extracts (", required_ng_PacBio," ng per extract) will have a volume of", total_uL, "µL.\n",
  "And a concentration of", total_ng, "ng /", total_uL, "µL =", pool_conc, "ng/µL.\n\n",
  "To reach a concentration of", target_conc, "ng/µL, we should condense the pool to", condensed_volume, "µL.\n"
)
```
## Export Selected Samples

This section prepares and exports the final list of selected samples with all relevant metadata for PacBio sequencing.

```{r pacbio_selection_export_05A, eval=FALSE, message=FALSE, fig.height=13, fig.width=20}
# Prepare export data frame with essential information
best_samples_export <- best_samples %>%
  select(EX_batch, extract, animal, tube_barcode, DNA_concentration, elution_uL, leftover_uL, leftover_ng, uL_required,
         microsample, animal_breed, treatment, age, richness, uL_left_after_pacbio, ng_left_after_pacbio)

best_samples_export

write.csv(best_samples_export, 
          file = "pacbio_selection_adenovirus.csv", row.names = FALSE)
```

## Save Data

This section saves the processed extract metadata and selected samples for use in other analyses (e.g., LMD selection).

```{r wrap_objects_05A, eval=FALSE}
# Save extract concentration data and selected samples
# This file is used in subsequent selection analyses (LMD, Metatranscriptomics)
save(extract_sample_div_metadata_macro, best_samples, best_samples_export,
  file = "data/macro/extract_concentration_data.Rdata"
)
```
