# LMD Selection Analysis

## Overview

This chapter identifies samples suitable for Laser Microdissection (LMD) based on DNA concentration and alpha diversity metrics. LMD requires high-quality DNA samples with sufficient quantity and taxonomic diversity for downstream spatial analyses. This analysis prioritizes samples with high richness and adequate DNA yield from day 7 post-infection samples.

## Load Data

```{r load_data_05B, eval=TRUE, message=FALSE, warning=FALSE}
# Load data required for plotting
# Note: extract_concentration_data.Rdata is created in 05A-PacBio_selection.Rmd
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
load("data/macro/extract_concentration_data.Rdata")
```

## Identify Best LMD Samples

This section identifies the best samples for LMD by ranking all day 7 samples based on combined richness and DNA concentration metrics.

```{r lmd_selection_all_05B, eval=TRUE, message=FALSE, fig.height=15, fig.width=18}
# Select best LMD samples based on combined ranking of richness and DNA concentration
# Filter for day 7 samples only (optimal time point for LMD analysis)
# Rank by combined score: richness (weighted 1.2x) + DNA concentration
# Select top 2 samples per breed × treatment combination
best_samples_LMD <- extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  filter(age == 7) %>%
  group_by(animal_breed, treatment_expl, age) %>%
  mutate(
    richness_rank = rank(desc(richness), ties.method = "min"),
    leftover_rank = rank(desc(DNA_concentration), ties.method = "min"),
    combined_rank = richness_rank*1.2 + leftover_rank
  ) %>%
  slice_min(combined_rank, n = 2, with_ties = FALSE) %>%
  ungroup()
```

## Broiler Sample Selection

This section identifies and visualizes the best LMD samples for broilers based on combined ranking of richness and DNA concentration.

### Visualize Broiler Samples

This section creates visualizations to assess broiler samples for LMD selection, showing richness, DNA concentration, and taxonomic composition.

```{r lmd_sample_selection_broiler_05B, eval=TRUE, message=FALSE, comment=""}
# Plot 1: Broiler richness by animal, colored by DNA concentration
# Shows alpha diversity combined with DNA quality metrics
barplot_2 <- extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "broiler") %>%
  ggplot(aes(x = animal, y = richness, fill = DNA_concentration)) +   # leftover_ng
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "DNA extracts", y = "richness", fill = "Read type") +
  scale_fill_gradient(low = "#c90076", high = "#3598bf", name = "extract DNA concentration") +
  facet_nested(. ~ animal_breed + treatment_expl + treatment + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(), #axis.text.x = element_blank(),
        legend.position = "top",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers")
# barplot_2


# Plot 2: Broiler taxonomic composition by order
# Shows community structure to ensure taxonomic diversity in selected samples
barplot_3 <- tidy_plot_genome_counts_filt_30_macro %>%
  filter(animal_breed == "broiler") %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal , y = count, fill = order)) +
  geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
  scale_fill_manual(values = order_colors[-4], drop = FALSE) +
  labs(x = "Composition",y = "Microsamples",fill = "Read type") +
  facet_nested(. ~ animal_breed + treatment_expl + treatment + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_blank(),
        legend.position = "none") +
  ggtitle("Composition of bacterial orders - 30% filt")
# barplot_3

best_samples_LMD_broiler <- best_samples_LMD %>%
  filter(animal_breed == 'broiler')
```


### Mark Selected Broiler Samples

This section marks the selected broiler samples on the visualization plots.

```{r lmd_barplots_broiler_marked_05B, eval=TRUE, message=FALSE, fig.height=13, fig.width=20}
# Mark selected broiler samples with star symbols on the richness plot
barplot_4 <- barplot_2 +
  geom_point(
    data = best_samples_LMD_broiler,
    aes(x = animal, y = richness),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

barplot_5 <- barplot_3 +
  geom_point(
    data = best_samples_LMD_broiler,
    aes(x = animal, y = 1),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

# barplot_4/barplot_5

```

### PCA Analysis for Broilers

This section performs PCA analysis on broiler samples to visualize community structure and highlight selected samples in ordination space.

```{r lmd_pca_broiler_05B, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Prepare community matrix for PCA analysis
# Filter for broiler samples at day 7, convert to wide format, and normalize to percentages
PCA_df <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(animal_breed == "broiler") %>%
  filter(age == 7) %>%
  # Remove metadata
  select(genome, microsample, count) %>%
  # Make tidy df into wide df
  pivot_wider(
    names_from = genome,   
    values_from = count,   
    # Fill missing values with 0
    values_fill = 0) %>%
  # Remove taxa (columns) that are zero for all leftover samples (rows)
  select(where(~ any(. != 0))) %>%
  # scale each sample to 100 (= closure to 100, not required, but common)
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100 )) %>%
  # turn the column 'microsample'into the name of the rows of the df
  column_to_rownames(var = "microsample")


# Create sample metadata with selection status for visualization
MACRO_sample_metadata <- sample_metadata_macro %>%
  mutate(microsample = sample) %>%
  select(microsample, animal, animal_breed, age, treatment, treatment_expl) %>%
  mutate(selection_status = if_else(animal %in% best_samples_LMD_broiler$animal, 
                                    "selected", 
                                    "not-selected"))



# Perform PCA with CLR transformation
pca_result <- perform_pca(PCA_df, z_delete = TRUE)

# Create PCA ordination plot
# Colors indicate treatment, shapes indicate selection status (selected vs not-selected)
p <- plot_pca(pca_result$pca_result,
         samples_color_metadata = "treatment" , 
         samples_shape_metadata = "selection_status",
         samples_color_value = treatment_colours_bright, 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         show_labels = TRUE, 
         MACRO_sample_metadata, genome_metadata, order_colors, 
         custom_ggplot_theme)

p2 = p 
# p2  # Individual plot commented out - used in combined plot

# # Calculate Euclidean distance on CLR data
# pca_animals_dist_matrix <- vegdist(pca_result$df_clr_dist, method = "euclidean")
# pca_animals_dist_matrix

```


```{r lmd_combined_broiler_plot_05B, eval=TRUE, message=FALSE, fig.height=11, fig.width=19}
# Combined visualization: barplots and PCA for broilers
# Shows richness, composition, and ordination space with selected samples highlighted
(barplot_4/barplot_5) | p2
```


## Layer Sample Selection

This section identifies and visualizes the best LMD samples for layers based on combined ranking of richness and DNA concentration.

### Visualize Layer Samples

This section creates visualizations to assess layer samples for LMD selection, showing richness, DNA concentration, and taxonomic composition.

```{r lmd_sample_selection_layer_05B, eval=TRUE, message=FALSE, comment=""}
# Plot 1: Layer richness by animal, colored by DNA concentration
# Shows alpha diversity combined with DNA quality metrics
barplot_2 <- extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "layer") %>%
  ggplot(aes(x = animal, y = richness, fill = DNA_concentration)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "DNA extracts", y = "richness", fill = "Read type") +
  scale_fill_gradient(low = "#c90076", high = "#3598bf", name = "extract DNA concentration") +
  facet_nested(. ~ animal_breed + treatment_expl + treatment + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(),
        legend.position = "top",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Layers")
# barplot_2

# Plot 2: Layer taxonomic composition by order
# Shows community structure to ensure taxonomic diversity in selected samples
barplot_3 <- tidy_plot_genome_counts_filt_30_macro %>%
  filter(animal_breed == "layer") %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal , y = count, fill = order)) +
  geom_bar(stat="identity", colour="white", linewidth=0.1) +
  scale_fill_manual(values = order_colors[-4], drop = FALSE) +
  labs(x = "Composition",y = "Microsamples",fill = "Read type") +
  facet_nested(. ~ animal_breed + treatment_expl + treatment + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_blank(),
        legend.position = "none") +
  ggtitle("Composition of bacterial orders - 30% filt")
# barplot_3

# Note: best_samples_LMD was already created in lmd_selection_all_05B chunk
# This section just creates the layer-specific visualizations
```


```{r lmd_filter_layer_05B, eval=TRUE, message=FALSE, fig.height=15, fig.width=18}
best_samples_LMD_layer <- best_samples_LMD %>%
  filter(animal_breed == 'layer')
```


### Mark Selected Layer Samples

This section marks the selected layer samples on the visualization plots.

```{r lmd_barplots_layer_marked_05B, eval=TRUE, message=FALSE, fig.height=13, fig.width=20}
# Mark selected layer samples with star symbols on the richness plot
barplot_4 <- barplot_2 +
  geom_point(
    data = best_samples_LMD_layer,
    aes(x = animal, y = richness),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

barplot_5 <- barplot_3 +
  geom_point(
    data = best_samples_LMD_layer,
    aes(x = animal, y = 1),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

# barplot_4/barplot_5

```

### PCA Analysis for Layers

This section performs PCA analysis on layer samples to visualize community structure and highlight selected samples in ordination space.

```{r lmd_pca_layer_05B, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# Prepare community matrix for PCA analysis
# Filter for layer samples at day 7, convert to wide format, and normalize to percentages
PCA_df <- tidy_plot_genome_counts_filt_30_macro_closed %>%
  filter(animal_breed == "layer") %>%
  filter(age == 7) %>%
  # Remove metadata
  select(genome, microsample, count) %>%
  # Make tidy df into wide df
  pivot_wider(
    names_from = genome,   
    values_from = count,   
    # Fill missing values with 0
    values_fill = 0) %>%
  # Remove taxa (columns) that are zero for all leftover samples (rows)
  select(where(~ any(. != 0))) %>%
  # scale each sample to 100 (= closure to 100, not required, but common)
  mutate(across(-microsample, ~ . / rowSums(across(where(is.numeric))) * 100 )) %>%
  # turn the column 'microsample'into the name of the rows of the df
  column_to_rownames(var = "microsample")


# Create sample metadata with selection status for visualization
MACRO_sample_metadata <- sample_metadata_macro %>%
  mutate(microsample = sample) %>%
  select(microsample, animal, animal_breed, age, treatment, treatment_expl) %>%
  mutate(selection_status = if_else(animal %in% best_samples_LMD_layer$animal, 
                                    "selected", 
                                    "not-selected"))



# Perform PCA with CLR transformation
pca_result <- perform_pca(PCA_df, z_delete = TRUE)

# Create PCA ordination plot
# Colors indicate treatment, shapes indicate selection status (selected vs not-selected)
p <- plot_pca(pca_result$pca_result,
         samples_color_metadata = "treatment" , 
         samples_shape_metadata = "selection_status",
         samples_color_value = treatment_colours_bright, 
         loadings_color_metadata = "order", 
         loadings_color_value = order_colors, 
         loadings_taxon_level = "species",
         show_labels = TRUE, 
         MACRO_sample_metadata, genome_metadata, order_colors, 
         custom_ggplot_theme)

p2 = p 
# p2  # Individual plot commented out - used in combined plot

# # Calculate Euclidean distance on CLR data
# pca_animals_dist_matrix <- vegdist(pca_result$df_clr_dist, method = "euclidean")
# pca_animals_dist_matrix

```

```{r lmd_combined_layer_05B, eval=TRUE, message=FALSE, fig.height=11, fig.width=19}
# Combined visualization: barplots and PCA for layers
# Shows richness, composition, and ordination space with selected samples highlighted
(barplot_4/barplot_5) | p2 
```


## Export Selected Samples

This section prepares and exports the final list of selected LMD samples.

```{r lmd_selection_export_05B, eval=FALSE, message=FALSE, fig.height=13, fig.width=20}
# Prepare export data frame with essential information
best_samples_LMD_export <- best_samples_LMD %>%
  select(animal, animal_breed, treatment, age, richness, microsample)

best_samples_LMD_export

write.csv(best_samples_LMD_export, 
          file = "LMD_selection_adenovirus.csv", row.names = FALSE)
```

