# Metatranscriptomics Selection Analysis

## Overview

This chapter identifies samples suitable for metatranscriptomics sequencing based on RNA concentration, elution volume, and taxonomic diversity metrics. Metatranscriptomics requires high-quality RNA samples with sufficient quantity and concentration for library preparation, typically targeting samples with high RNA yield and taxonomic diversity.

## Load Data

```{r load_data_05C, eval=TRUE, message=FALSE, warning=FALSE}
# Load processed plotting data, color schemes, and extract concentration data
load("data/macro/plot_data.Rdata")
load("data/data_colors.Rdata")
load("data/macro/extract_concentration_data.Rdata")
```

## Prepare RNA Extract Metadata

This section loads RNA extraction metadata from Airtable, including concentration, elution volume, and spatial coordinates. This metadata is used to identify samples meeting metatranscriptomics requirements.

```{r load_rna_extract_metadata_05C, eval=FALSE, message=FALSE}
# Load RNA extraction metadata from Airtable
# This includes concentration, elution volume, and spatial coordinates (SU_coordinate) for each extract
RNA_extract_metadata_macro <- airtable("2-EX-Info", "apphvsV2YKhTCMKGS") %>% 
  # select columns to extract
  read_airtable(., fields = c(
    "Extract", "Barcode", "Concentration_ng_uL", "ElutionVolume_uL",
    "LeftOver_uL", "LeftOver_ng", "AcidNucleicMaterial",
    "LabBatch_text", "ExperimentalUnit", "SUcoordinate_text"), id_to_col = TRUE) %>%
  # rename columns
  rename(
    EX_batch = "LabBatch_text",
    extract = "Extract",
    animal = "ExperimentalUnit",
    tube_barcode = "Barcode",
    RNA_concentration = "Concentration_ng_uL",
    elution_uL = "ElutionVolume_uL",
    leftover_uL = "LeftOver_uL",
    leftover_ng = "LeftOver_ng",
    nucleic_acid = "AcidNucleicMaterial",
    SU_coordinate = "SUcoordinate_text"
  ) %>%
    # select relevant batches
  filter(EX_batch %in% c("EXB020","EXB021")) %>%
  unnest(c(nucleic_acid, animal, SU_coordinate)) %>%
  # select relevant columns
  select(EX_batch, extract, animal, tube_barcode, RNA_concentration, elution_uL, nucleic_acid, SU_coordinate) %>%
  filter(nucleic_acid == "RNA") %>%
  # un-nest columns that appeared as a list
  arrange(animal)
```      


## Merge RNA Metadata with Sample Information

This section merges RNA extract metadata with sample metadata and alpha diversity metrics to create a comprehensive dataset for sample selection.

```{r merge_rna_extract_metadata_05C, eval=FALSE, message=FALSE}
# Merge RNA extract metadata with sample metadata
# This links RNA extraction information to experimental samples (animal, treatment, breed, age)
RNA_extract_sample_metadata_macro <- RNA_extract_metadata_macro %>% 
  left_join(sample_metadata_macro, by = join_by(animal == animal)) %>%
  rename(microsample=sample)

RNA_extract_sample_div_metadata_macro <- RNA_extract_sample_metadata_macro %>%
 left_join(alpha_div_macro_filtered_30, by = join_by(microsample == microsample)) 
```


## Visualize RNA Concentration and Richness

This section creates visualizations to assess RNA extraction quality and taxonomic richness across all samples, helping identify candidates for metatranscriptomics sequencing.

```{r rna_barplots_richness_05C, eval=TRUE, message=FALSE, fig.height=6, fig.width=15}
# Plot 1: Taxonomic richness by animal, colored by RNA concentration
# Shows alpha diversity combined with RNA quality metrics
barplot_RNA2   <- RNA_extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal, y = richness, fill = RNA_concentration)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "RNA extracts", y = "richness", fill = "Read type") +
  scale_fill_gradient(low = "#c90076", high = "#3598bf", name = "RNA concentration") +
  facet_nested(. ~ animal_breed + treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(), #axis.text.x = element_blank(),
        legend.position = "top",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers + Layers")
# barplot_RNA2  # Individual plot commented out - used in combined plot
```


```{r rna_barplots_conc_05C, eval=TRUE, message=FALSE, fig.height=6, fig.width=15}
# Plot 2: RNA concentration by animal, colored by taxonomic richness
# Shows RNA quality metrics combined with diversity information
barplot_RNA2 <- RNA_extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  ggplot(aes(x = animal, y = RNA_concentration, fill =  richness)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "RNA extracts", y = "RNA concentration", fill = "Read type") +
  scale_fill_gradient(low = "#c90076", high = "#3598bf", name = "richness") +
  facet_nested(. ~ animal_breed + treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(), #axis.text.x = element_blank(),
        legend.position = "top",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle("Broilers + Layers")
# barplot_RNA2  # Individual plot commented out - used in combined plot
```



## Identify Best Samples

This section applies selection criteria to identify the best samples for metatranscriptomics sequencing, ranking samples by combined RNA concentration and richness metrics.

```{r rna_selection_prep_05C, eval=TRUE, message=FALSE, fig.height=6, fig.width=15}
# Step 1: Filter samples based on quality criteria
# Requirements: RNA concentration > 30 ng/uL, non-missing values, specific time points
rna_filtered <- RNA_extract_sample_div_metadata_macro %>%
  filter(
    !is.na(RNA_concentration),
    !is.na(richness),
    RNA_concentration > 30,
    (animal_breed == "broiler" & age %in% c(5, 14) |
     animal_breed == "layer"   & age %in% c(7, 14) |
     age == 0)  # optional: include day 0 controls
  )

# Step 2: Rank samples by combined score
# Normalize metrics to 0-1 scale and create weighted combined score
# Weight: 60% RNA concentration, 40% richness (prioritizes RNA quality)
rna_ranked <- rna_filtered %>%
  mutate(
    # Normalize richness and RNA concentration (0–1 scaling) for fair comparison
    richness_scaled = scales::rescale(richness),
    RNA_scaled = scales::rescale(RNA_concentration),
    # Create a combined score (weighted average favoring RNA quality)
    combined_score = 0.6 * RNA_scaled + 0.4 * richness_scaled
  ) %>%
  arrange(desc(combined_score))  # Sort best to worst

# Step 3: Select top samples per group
# Select top 3 samples per treatment × age combination for broilers
# Note: This can be modified to include layers or other grouping schemes
top_per_group <- rna_ranked %>%
  filter(animal_breed=="broiler") %>%
  group_by(treatment, age) %>%
  slice_max(order_by = combined_score, n = 3, with_ties = FALSE) %>%
  ungroup()
```




## Visualize Selected Samples

This section marks selected samples on the visualization plots to show which samples were chosen for metatranscriptomics sequencing.

```{r rna_barplots_marked_05C, eval=TRUE, message=FALSE, fig.height=6, fig.width=13}
# Create barplot showing RNA concentration by animal, colored by richness
# Filtered to show only broiler samples
barplot_RNA3 <- RNA_extract_sample_div_metadata_macro %>%
  filter(treatment != "TF0") %>%
  filter(animal_breed == "broiler") %>%
  ggplot(aes(x = animal, y = RNA_concentration, fill =  richness)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  labs(x = "RNA extracts", y = "RNA concentration", fill = "Read type") +
  scale_fill_gradient(low = "#c90076", high = "#3598bf", name = "richness") +
  facet_nested(. ~ treatment_expl + age, scales = "free", space = "free") +
  custom_ggplot_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(), #axis.text.x = element_blank(),
        legend.position = "top",
        plot.margin = margin(0.2, 0.2, 0.2, 0.2)) +
  ggtitle(" broiler")


# Mark selected samples with star symbols
barplot_RNA2_3 <- barplot_RNA3 +
  geom_point(
    data = top_per_group,
    aes(x = animal, y = RNA_concentration),
    shape = 8,          # star shape
    size = 4,
    color = "black",
    inherit.aes = FALSE
  )

# Display plot with selected samples marked
barplot_RNA2_3  
```

## Export Selected Samples

This section prepares and exports the final list of selected samples with all relevant metadata for metatranscriptomics sequencing.

**Novogene sequencing requirements:**
- Total RNA: ≥300 ng
- Volume: ≥20 μL
- Concentration: ≥15 ng/μL

```{r rna_selection_export_05C, eval=FALSE}
# Define shipping volume per sample (22 uL meets Novogene requirements)
ship_to_Novogene_volume_uL_value = 22

# Prepare export data frame with essential information
# Calculate shipping quantities and remaining volumes
best_RNA_samples_export <- top_per_group %>%
  select(nucleic_acid, EX_batch, extract, animal, SU_coordinate, tube_barcode, RNA_concentration, elution_uL) %>%
  mutate(
    ship_to_Novogene_volume_uL = ship_to_Novogene_volume_uL_value,  # Volume to ship
    ship_ng = ship_to_Novogene_volume_uL*RNA_concentration,        # Quantity to ship
    leftover_uL = elution_uL - ship_to_Novogene_volume_uL          # Remaining volume
  )

write.csv(best_RNA_samples_export, 
          file = "transcriptomics_selection_adenovirus.csv", row.names = FALSE)
```


## Save Data

This section saves the processed RNA extract metadata and selected samples for future reference.

```{r wrap_objects_05C, eval=FALSE}
# Save RNA extract concentration data and selected samples
save(RNA_extract_sample_div_metadata_macro, top_per_group, rna_ranked, best_RNA_samples_export,
  file = "data/macro/RNA_extract_concentration_data.Rdata"
)
```

