---
title: "3D'omics | Adenovirus: Conventional metagenomics"
author:
- Amalia Bogri^[University of Copenhagen, amalia.bogri@sund.ku.dk]
- Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
- Jorge Langa^[University of Copenhagen, jorge.langa@sund.ku.dk]

date: "`r Sys.Date()`"
subtitle: Work in progress - preliminary analysis
site: bookdown::bookdown_site
documentclass: book
# bibliography:
# - book.bib
# - packages.bib
url: "https://github.com/3d-omics/AMAC009-adenovirus_chicken.git"
description: |
  Data analysis code for 3D'omics adenovirus trial using conventional metagenomics
link-citations: true
github-repo: "3d-omics/XXX"
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
  class.source = "script-source",
  class.output = "script-output",
  comment = NA
)
```

# Introduction

This webbook contains all the code used for adenovirus trial analysis that is presented in EU 3D'omics Deliverable 6.3.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/3d-omics/AMAC009-adenovirus_chicken.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(devtools)
library(tinytable)
library(broom)
library(broom.mixed)
library(rairtable)
library(jsonlite)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)
library(jpeg)
library(patchwork)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(lme4)
library(Hmsc)
library(psych)

# For compositional analyses
library(zCompositions)
library(compositions)
library(ALDEx2)
library(ANCOMBC)
library(microbiome)
library(ggcorrplot)
library(propr)
library(ggpmisc)
library(ade4)
library(spdep)

library(janitor)
library(permuco)
library(magick)

library(MASS) # for Negative Binomial Fit

library(ggVennDiagram)

library(Hmisc)
library(reshape2)
# Tidyverse libraries (load dplyr last to avoid conflicts)
library(tidyverse)
library(dplyr) # Explicitly load dplyr last
```

## Load Essential Functions

The following functions are required for PCA and other compositional analyses. These are loaded early to ensure they are available throughout the bookdown.

```{r load_pca_functions, warning=FALSE, message=FALSE}
# Source PCA and compositional analysis functions
source("pca_functions.R")
```

## Notes
### Installation of modules
Most of the modules can be downloaded from CRAN with 'install.packages('XXX', dependencies = TRUE)'.

Four modules are from Bioconductor. 
Install with: 'install.packages('BiocManager', dependencies = TRUE)
BiocManager::install(c("phyloseq", "ggtreeExtra","ggtree","ANCOMBC"))'

Two modules are from Antton's github. 
Install with: 'install.packages('remotes', dependencies = TRUE)
remotes::install_github('anttonalberdi/hilldiv2')
remotes::install_github('anttonalberdi/distillR')'

if (!require("BiocManager", quietly = TRUE))
     install.packages("BiocManager")
BiocManager::install("ALDEx2")
remotes::install_github('tpq/propr')


### Airtable access:
Currently the metadata is loaded from the 3D'omics airtable. 
In order to access it you need your own API key (from Antton ALberdi). 
Use set_airtable_api_key('XXXXX', install = TRUE).

